---
title: "Week 03 - Map Classification"
author: "Adela Sobotkova"
date: "06/01/2021 updated`r Sys.Date()`"
output:
  rmdformats::readthedown:
    highlight: kate
---

```{r setup, include=FALSE}
library(knitr)
library(rmdformats)

## Global options
options(max.print="75")
opts_chunk$set(echo=TRUE,
	             cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```

##  Learn about Classification methods

This exercise demonstrates some possibilities of classifications in R, based on Ch.9 of [Geocomputation](https://geocompr.robinlovelace.net/adv-map.html)

### Required R libraries

* We will use the sf, raster, and tmap packages.
* Additionally, we will use the spData and spDataLarge packages that provide new datasets.

```{r}
library(sf)
library(raster)
library(tmap)
library(spData)
library(spDataLarge)
```


### Data sets 

We will use a single data set:  `nz`. It is contained by the libraries
It is an sf object with polygons representing the 16 regions of New Zealand.

### Existing code

Here are some examples of plotting population in  New Zealand. * Your role is to create a map based on the suggestions below, selecting the most meaningful classification style.

```{r nz-pop}
# Look at NZ population distribution
hist(nz$Population)
```


```{r maps}
# This line of code applies the 'pretty' style rounding legend numbers. Try different numbers of classes.
a <- tm_shape(nz) + tm_polygons(col = "Population", style = "pretty", n = 4)

# "Jenks" style further smooths over the gaps
b <- tm_shape(nz) + tm_polygons(col = "Population", style = "jenks", n = 5)

# quantile style divides into 5 even groups
c <- tm_shape(nz) + tm_polygons(col = "Population", style = "quantile", n=5)

# Equal interval style divides the distribution into even groups
d <- tm_shape(nz) + tm_polygons(col = "Population", style = "equal", n = 5)

# Write maps above to objects and plot them side by side 
# with tmap_arrange() for better comparison
tmap_arrange( ___,___, ___,___)
```


#### Questions

* 1. What are the advantages and disadvantages of each classification method?
* 2. Choose the best classification and create a map with easily legible legend and all other essentials.(Select a suitable color palette from http://colorbrewer2.org/, north arrow, scale, map title, legend title, reasonable number of breaks in the classification )
* 3. Which method and how many classes did you end up using to display your data? Why did you select that method?
* 4. What principles did you use in arranging the parts of your map layout the way you did and how did you apply these principles?

# Your solution

```{r}
# /Start Code/ #
pdf("nzmap.pdf")
tmap_arrange(a,b,c,d)
dev.off()

# /End Code/ #
```

