---
title: "W03b - Bounding boxes, Buffers and centroids"
author: "Adela Sobotkova"
date: "06/01/2021 updated`r Sys.Date()`"
output:
  rmdformats::readthedown:
    highlight: kate
---

```{r setup, include=FALSE}
library(knitr)
library(rmdformats)

## Global options
options(max.print="75")
opts_chunk$set(echo=TRUE,
	             cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```

# Task 1: Create spatial data and buffer layers

Computing buffers is a key spatial analysis skill and the resulting buffers have a wide range of uses like, for example, identifying the number of roads within one kilometer of a school or computing the number of hazardous waste sites near sensitive natural areas.

Although, technically you can buffer data with unprojected coordinate reference systems, the buffer distance will be more meaningful with a projected CRS so it is highly recommended that you transform unprojected data to a projected CRS before buffering.

## Instructions

* Load the packages `sf` and `raster`along with the objects `df`, `shelters`, and `aarhus` in your workspace.
* Use `st_as_sf()` to convert this data frame to an `sf` object, use 4326 for the `crs` argument because these points are latitude/longitude -- an unprojected CRS.
* Use `st_transform()` to transform the points' CRS to match the `shelters`  so we can plot them together. You will need the `st_crs()` function to get the CRS of the `shelters` or use the EPSG 25832.
* Buffer both the `shelters` and your points by 500 meters (no need to specify "meters", since this is the projection unit, this will be the default). The radius argument is `dist`.
* Plot the `shelters`, shelter buffers,  `aarhus` image, and the geometries of buffers and points together. This will require running `plot()`,`plot()`, `plotRGB()`, and `plot()` again, all together.

```{r buffer, eval=FALSE}
# Create dataframe with lat long data
df <- data.frame(place = c("Aarhus Domkirke", 
                           "Christianskirken",
                           "Nobelparken"),
                 longitude=c(10.2102364,10.1949197,10.2033485),
                 latitude=c(56.1574919,56.1772228,56.174346))


# Convert the data frame to an sf object             
df_sf <- ___(df, coords = c("longitude", "latitude"), ___)

# Plot the kaz image (it is multi-band)
___(kaz)
___(___(df_buf), col = "firebrick", add = TRUE)
___(___(df_crs), pch = 16, add = TRUE)


# Transform the points to match the shelters' CRS EPSG 25832
df_crs <- st_transform(df_sf, crs = 25832)

# Buffer the points
df_buf <- _____(df_crs, dist = 500)


# Load the aarhus image


# Load shelters and buffer them by 500m
shelters <- 
shelters_buf <- _____(shelters, _____)

# Plot the shelters and buffers, aarhus image and points and their buffers
_____(______(shelters_buf), col = "green")
_____(_______(shelters), pch = 2 ,add = TRUE)
_____(aarhus, stretch = "lin",add = TRUE)
_____(______(df_buf), col = "firebrick",add = TRUE)
_____(______(df_crs), pch = 16, add = TRUE)

```


Congratulations! Buffering is pretty straightforward in `sf`. The hardest part is probably getting your data into R and ensuring that they have the right coordinate reference system. The map is far from splendid, but having all the objects there is a success :)

# Task 2: Compute polygon centroids
Similar to buffering, computing polygon centroids is a bedrock geoprocessing task used to assign values and even to help with labeling maps. The function for this in sf is `st_centroid()`.

Also similar to buffering, centroid calculations should generally be performed on data with a projected coordinate reference system.

## Instructions

* Read in the `municipalities` shapefile from GADM database (see Bonus in Week 02).
* Project/transform the `municipalities` using the EPSG number 25832 as the target CRS. As the municipalities is a SpatialPolygons Dataframe, you will need to use `st_as_sf()` to convert it to `sf` object first.
* Compute the centroids with `st_centroid()` and store as an object named `centroids`. Only one argument is required.
* Plot the geometry of `municipalities` object with the centroids on top. This requires two calls to `plot()` run together. The second will need `add = TRUE`.

```{r centroids, eval=FALSE}
# Read in the municipality shapefile
municipalities <- ___

# Convert municipalities to sf object and project to epsg 25832
municipalities_tf <- ___(__________(municipalities), crs = 32618)

# Compute the municipality centroids
centroids <- ___(municipalities_tf)

# Plot the municipality geometry
plot(___(municipalities_tf), col = "grey", border = "white")
plot(centroids, pch = 16, col = "firebrick", ___ = ___)
```


Complete! You will find that knowing how to compute polygon centroids will serve you well in many contexts, from attaching labels to polygons in R to calculating neighborhoods and distances.

# Task 3: Create a bounding box around vector data
You can compute bounding boxes around vector data using `sf`. These can help you, for example, create polygons to clip layers to a common area for an analysis (e.g to limit area to aarhus municipality) or identify regions of influence.

In the `sf` package, there is a function for extracting the bounding box coordinates, if that's all you need, this is `st_bbox()`. More likely you'll want to create a new `sf` object (a polygon) from those coordinates and to do this `sf` provides the `st_make_grid()` function.

`st_make_grid()` can be used to make a multi-row and multi-column grid covering your input data but it can also be used to make a grid of just one cell (a bounding box). To do this, you need to specify the number of grid cells as `n = 1`.

## Instructions

* Filter out Aarhus kommune from the `municipalities_tf` sf object and create a new `aarhus_mun` object. How exactly is Aarhus spelled and which column do you use? 
* Review that your selection worked by plotting `shelters` over the `aarhus_mun` object - they should overlap.
* Compute the bounding box coordinates with `st_bbox()`. What is the result?
* Create a single polygon bounding box around the aarhus municipality using `st_make_grid()` with an argument of `n = 1`.
* Create a plot of the `municipalities`, highlighting the aarhus municipality and `aarhus_box` (just the geometry), the new box on top of it. 
* Create a second plot with the `aarhus_box` and `shelters` on top of the aarhus ortophoto.

```{r Ar-box, eval= FALSE}
# Filter out the Aarhus kommune from the municipalities sf object
aarhus_mun <- _______________ %>% 
  filter(____________)

# Plot the aarhus municipality and the shelters
plot(_______(aarhus_mun), col = "grey", border = "white")
plot(________, add = TRUE, pch = 16, col = "forestgreen")

# Compute the coordinates of the bounding box
_________(aarhus_mun)

# Create a bounding box polygon
aarhus_box <- __________(aarhus_mun, n = ______)

# Plot all the DK municipalities, add the shelters and the new box outlining the Aarhus municipality
plot(st_geometry(municipalities_tf), col = "grey", border = "white")
plot(________, _________, pch = 16, col = "forestgreen")
plot(________, _________)
```


Bounding box completed! You can see in the plot that the bounding box is the minimum rectangular polygon that fits all the shelters. You might now clip a series of layers to this box for a consistent extent over the Aarhus municipality. You will learn how to do this in a later exercise.


# Task 4: Dissolve multiple features into one
In order to compute a tighter bounding box, **a convex hull**, around a set of points like the shelters or other features, which do not neatly fit a rectangular shape you'll need to learn one more function first.

For points, remember and beware that the convex hull function creates a boundary around **each** point! This is unlikely what you want. More likely you want to compute a convex hull around all your points. If you have a set of points and you want to draw a convex hull around them you first need to bundle the points into a single MULTIPOINT feature and in order to do this you will use the dissolve function in `sf` called `st_union()`.

With polygons, `st_union()` will dissolve all the polygons into a single polygon representing the area where all the polygons overlap. Your set of individual points will be dissolved/unioned into a single, MULTIPOINT feature that you can use for tasks like computing the convex hull.

## Instructions

* Buffer the `shelters` object by 500m with `st_buffer()` -- `sf` will automatically use meters, the units of the CRS. Check the CRS of shelters and aarhus municipality to ensure they are identical and reproject if not. Use pipes for neater code.
* Create a new object called `sh_buffers` which is just the geometry of the buffered shelters with `st_geometry()`.
* Compute the number of features in the `sh_buffers` object with `length()` and `plot()` to see what they look like.
* Dissolve the buffers in `sh_buffers`, call this `sh_buf_union`.
* Compute the number of features in the `sh_buf_union` object with `length()` and `plot()` to see what the dissolved object looks like.

```{r shelter-buffers, eval=FALSE}
# Buffer the shelters by 500m after projecting the object to crs
all.equal(___________,________)

sh_buffer <- shelters %>% 
  # project to Aarhus municipality CRS if different
  ___________(crs = ________) %>% 
  st_buffer(______)

# Limit the object to just geometry
sh_buffers <- _________(sh_buffer)

# Compute the number of features in sh_buffer
________(sh_buffers)

# Plot the shelter buffers
_____(sh_buffers)

# Dissolve the buffers
sh_buf_union <- __________(sh_buffers)

# Compute the number of features in sh_buf_union
________(sh_buf_union)

# Plot the dissolved buffers
______(sh_buf_union)

```


Great! You'll see later that the `st_union()` function is important in other spatial analysis contexts, so it's a good one to commit to memory.

# Task 5: Compute a convex hull around vectors
A more precise bounding polygon is sometimes needed, one that fits your data more neatly. For this, you can use the `st_convex_hull()` function. Note that `st_convex_hull()` will compute a tight box around each one of your features *individually* so if you want to create a convex hull around a group of features you'll need to use `st_union()` to combine individual features into a single multi-feature.

## Instructions

* Use `head()` on `shelters` to look at the data frame and see the type of geometry.
* Use `st_union()` to combine the individual points in the `shelters` object into a single MULTIPOINT geometry and call this sh1.
* Use `head()` on `sh1` to see the type of geometry of the dissolved object.
* Use the `length()` function from `base` R on `shelters` and `sh1` to confirm that the number of features went from 17 to 1.
* Use the `st_convex_hull()` function on `sh1` to compute the tight bounding box around the shelters and call this `shelters_hull`.
* Plot `shelters_hull` and then plot the points on top. Use `plot()` twice and run the lines together.

```{r convexhull, eval=FALSE}
# Look at the data frame to see the type of geometry
___(shelters)

# Convert the points to a single multi-point
sh1 <- ___(shelters)

# Look at the data frame to see the type of geometry
___(sh1)

# Confirm that we went from 17 features to 1 feature
___(shelters)
___(sh1)

# Compute the tight bounding box
shelters_hull <- ___(sh1)

# Plot the points together with the hull
plot(shelters_hull, col = "red")
plot(sh1, ___ = ___)
```


It takes a little getting used to, but when computing a convex hull, you'll need to remember to first *dissolve/union/combine* your individual features into a multi-feature. When you used `head()` you should have noticed that you started with “POINT” objects and after `st_union()` you had “MULTIPOINT”. Convex hulls will be super useful in later point analysis as statistical results often depend on the area that encloses the features of interest, and sometimes you need as tight as envelope as possible.


# Task 6: Spatial joins - Danish monuments
In this task, you will find out which municipality in Denmark has the most Viking-era monuments and finds registered. You will work with spatial data from [Fund og Fortidsminder](https://www.kulturarv.dk/fundogfortidsminder/Download/), namely a shapefile in ETRS89/UTM32N projection that collects information on 300,000 registered monuments.

For many analysis types you need to link geographies spatially. For example, you want to know how many Viking-era cultural monuments/artefacts (fortidsminder) are in each municipality but you don't have a municipality attribute in the data. The best way to do this is with a spatial join using `st_join()`.

Importantly, the `st_join()` function requires `sf` data frames as input and will not accept an object that is just `sf` geometry. You can use the `st_sf()` function to convert `sf` geometry objects to an sf data frame (`st_sf()` is essentially the opposite of `st_geometry()`).

## Instructions

* Download the fortidsminder archive from [Sciencedata](https://sciencedata.dk/shared/ce0f8e62af16dab66b45f13be90d00f8) and unzip. 
* Load the anlaeg_all_25832.shp into a `monuments` sf object. It might take couple seconds, because there are 300k of them.
* Filter the monuments object to those from the Viking-era ("Vikingtid")
* Plot the monuments (`viking`) on top of the municipalities. You will want to plot only the geometry of the municipalities.
* Use `class()` to see if the `viking` object has class `data.frame` or if it's just geometry.
* Use `st_join()` to conduct a spatial join in order to add municipality information to the `viking` object.
* Use `head()` to confirm that the new object has municipality information -- for example, it should now have municipality name (`NAME_2`).
* What five municipalities contain the most monuments? Which have the least and why? 
* Challenge: can you plot those five municipalities that contain the fewest viking finds on a map of Denmark? You can use the `st_intersect()` function and select municipalities which satisfy the intersection with fewest finds. 

```{r monuments-munic, eval=FALSE}
# Load the monuments into R
monuments <- read_sf(" _________")

# Check the periodisation and type of the monuments
_____(monuments$datering)
_____(monuments$anlaegsbet)

# Filter Viking-era finds
viking <- monuments %>% 
  _____(___________)

# Plot the monuments on top of the municipalities
plot(_________(municipalities_tf))
plot(_________, add = TRUE, pch = 16, col = "red")


# Determine whether viking monuments has class data.frame
class(viking)

# Join the viking monuments with the municipalities
viking_neigh <- ________(viking, municipalities_tf)

# Confirm that viking_neigh has the municipality information
head(viking_neigh)

# Calculate the count of monuments per municipality
viking_neigh %>% 
  group_by(________) %>% 
  tally() %>% 
  arrange(____(__)) 

# Create an object with 5 of the least populated municipalities
least <- viking_neigh %>% 
  group_by(________) %>% 
  tally() %>% 
  arrange(__) %>%
  _____(1:5)
  
  
# Plot the municipalities with fewest Viking finds in red over DK municipalities
plot(__________, col = "grey", border = "white")
plot(municipalities_tf[_________(___________(least, municipalities_tf, sparse = FALSE)), ]$geometry, col = "red", add = T) 

```

Great! The monuments object did not have municipality information until you performed the spatial join with `st_join()`. Now that the municipality information is included you can supply each municipality office with the count and coordinates of monuments they are in charge of so they can monitor and safeguard.

# Task 7: Spatial relationships
In the previous exercise, you loaded all monuments across Denmark. In this exercise you need to make a plan for family visit. Your family like to bike in green areas and visit viking monuments, so calculate:

*a) how many viking monuments are within 10000 m radius of Nobelparken and 
*b) how many parks are completely within this area as well.  

To achieve this, you will need to select the third object in the buffer (Nobelparken) and then use `st_intersects()` to tally the monuments within 10000m radius  and `st_contains()` to determine how many parks are *completely* within the buffer. You will then use the `st_intersection()` function (notice the slight difference in function name!) to **clip** the monuments and parks to the buffer.

A note about the output of functions that test relationships between two sets of features. The output of these and related functions is a special kind of list (with the class `sgbp`). For example, when using `st_intersects()`, the first element in the output can be accessed using `[[1]]`, which shows features from the monuments object that intersect with the buffer polygon. Likewise, `[[2]]` would show the buffer polygons that intersect with the monuments.

## Instructions

* Load `viking` and `parks` objects and verify consistent projection with `df_sf`
* Filter `df_sf` to Nobelparken only and create a 10000m buffer `nb_buf` around the point. There is something you will need to do before buffering!
* Use `st_intersects()` to identify `viking` that intersect with the buffer object (`nobel_buf`) and call the result `viking_int`.
* Use `st_contains()` to identify `parks` features that are completely within the buffer object (`buf`) and call the result `parks_cont`.
* Extract `viking` and parks that intersect and are contained by  `nobel_buf` and save as `int` and `cont`.
* Use the `int` object you just created to identify the IDs of the monuments that intersect with buffer (the first will be XXXX).
* Use `st_intersection()` to "clip" the monuments by `nobel_buf` and call this `viking_clip`. Do the same with parks. 
* Plot the result

```{r intersect, eval=FALSE}
# Load objects
parks
viking


# Create a 10km buffer around Nobelparken



# Check consistent projection among the objects
________==_________
________==_________

# Identify viking monuments that intersect with the buffer
viking_int <- ___(nb_buf, viking)

# Identify parks contained by the buffer
parks_cont <- ___(nb_buf, parks)

# Get the indexes of which viking monuments intersect and which units 
# are contained by the buffer
int <- viking_int[[___]]
cont <- parks_cont[[___]]

# Get the IDs or codes of the viking and parks in buffer
viking$TRAPCode[___]

# Clip the viking monuments and parks layers by the buffer (ignore the warning)
viking_clip <- ___
parks_clip <- ___

# Plot the geometry of the clipped viking monuments
plot(nb_buf$geometry[1,], border = "blue")
plot(parks[cont,], add = TRUE, col = "yellow")
plot(st_geometry(viking_clip), col = "red", add = TRUE, pch = 19)

```


#### Questions: 

1. What is the class of the object produced via `st_intersects()`?
2. Why do we use `parks[cont,]` to view the fully contained parks in the buffer plot?
3. How many parks and viking monuments respectively are there within the buffer?

The `sf` package has a lot of different functions for determining relationships. Some are easy to conflate!  You have just used `st_intersects()` and `st_contains()` here but there are others as well and what you use will depend on what relationships you're looking at. In this exercise you also used the important `st_intersection()` function to clip one layer based on another.

# Task 8: Measuring distance between features
Of course, measuring distance between feature sets is a component of spatial analysis 101 -- a core skill for any analyst. There are several functions in `base` R as well as in the packages `rgeos` and `geosphere` to compute distances, but the `st_distance()` function from `sf` provides a useful feature-to-feature distance matrix as output and can be used for most distance calculation needs.

In this exercise you'll measure the distance from the Nobelparken to the viking monuments within the 10km buffer and identify the 10 closest ones.

## Instructions

* Filter Nobelparken out of `db_sf` POINT feature and project to a planar CRS. 
* Test whether the CRS of `cities` matches the CRS of the preloaded object `viking_clip` with `st_crs()`.
* Use `st_distance()` to compute the distance between Nobelparken to the viking monuments within the 10km buffer .
* Use `head()` to take a quick look at the result.
* Identify the index of the 10 sites with the smallest distance. This may take several steps that include `sort()` and `which()` functions.
* Use `nearest10` to pull out the rows from `viking` of the nearest viking monument (see the `anlaegsbet` variable).

```{r nearest, eval=FALSE}
# Create the nobel object and transform to planar CRS
nobel <- ___(___)

# Test whether the CRS match
___(viking_clip) == ___(nobel)


# Compute the distance between Nobel and viking_clip
d <- ___(___, Kazanlak)


# Compute the distance between the Nobelparken and the clipped viking monuments
d <- ________(___________, nobel)

# Take a quick look at the result
________(d)

# Find the index of the 10 nearest viking monuments


# Identify the viking monuments that are the nearest, what kinds of monuments are they?
viking_clip[_________,]$anlaegsbet

# Plot the results by reruning the plot from exercise 6
```


Nice work! If you look at the result for the last line of code carefully you'll see that the closest viking monument is Befæstning at 1580m. In this case one of our feature sets was a single feature. You may end up applying this function in situations where there are multiple features in both objects. In this situation `sf` will return a matrix. Distance matrices take long time to calculate and so sometimes you calculate distance between the nearest two features from two different datasets. That will be the homework :)
