---
title: "W2- Exercise 1"
Author: "Adela Sobotkova and Peter Thramkrongart"
output: github_document
---


## Goals

Modify the provided code to improve the resulting map

We highlighted all parts of the R script in which you are supposed to add your
own code with: 

```{r}
print("Hello World") # This would be your code contribution
```



## Required R libraries 

We will use the sf, raster, and tmap packages.
Additionally, we will use the spData and spDataLarge packages that provide new datasets.
These packages have been preloaded to the worker2 workspace.
```{r}
library(sf)
library(raster)
library(tmap)
library(spData)
library(spDataLarge)
```

## Data sets

We will use two data sets: `nz_elev` and `nz`. They are contained by the libraries
The first one is an elevation raster object for the New Zealand area, and the second one is an sf object with polygons representing the 16 regions of New Zealand.

## Existing code

We wrote the code to create a new map of New Zealand.
Your role is to improve this map based on the suggestions below.

```{r}
tm_shape(nz_elev)  +
  tm_raster(title = "elev", 
            style = "cont",
            palette = "BuGn") +
  tm_shape(nz) +
  tm_borders(col = "red", 
             lwd = 3) +
  tm_scale_bar(breaks = c(0, 100, 200),
               text.size = 1) +
  tm_compass(position = c("LEFT", "center"),
             type = "rose", 
             size = 2) +
  tm_credits(text = "A. Sobotkova, 2020") +
  tm_layout(main.title = "My map",
            bg.color = "orange",
            inner.margins = c(0, 0, 0, 0))
```

## Exercise I

1. Change the map title from "My map" to "New Zealand".  
2. Update the map credits with your own name and today's date.  
3. Change the color palette to "-RdYlGn".   
    (You can also try other palettes from http://colorbrewer2.org/)  
4. Put the north arrow in the top right corner of the map.  
5. Improve the legend title by adding the used units (m asl).  
6. Increase the number of breaks in the scale bar.  
7. Change the borders' color of the New Zealand's regions to black.   
    Decrease the line width.  
8. Change the background color to any color of your choice.  

## Your solution

```{r}
tm_shape(nz_elev)  +
  tm_raster(title = "Elevation (m asl)", #5. Improve the legend title by adding the used units (m asl)
            style = "cont",
            palette = "-RdYlGn") + #3. Change the color palette to "-RdYlGn"
  tm_shape(nz) +
  tm_borders(col = "black", #7. Change the borders' color of the New Zealand's regions to black. Decrease the line width.  
             lwd = 1) +
  tm_scale_bar(breaks = c(0,50, 100,150, 200), #6. Increase the number of breaks in the scale bar
               text.size = 1) +
  tm_compass(position = c("RIGHT", "top"), #4. Put the north arrow in the top right corner of the map
             type = "rose", 
             size = 2) +
  tm_credits(text = "P. Thramkrongart, 2021") + #2. Update the map credits with your own name and today's date.  
  tm_layout(main.title = "New Zealand", #1. Change the map title from "My map" to "New Zealand"
            bg.color = "light blue", #8. Change the background color to any color of your choice. 
            inner.margins = c(0, 0, 0, 0))
```


##Exercise II

9. Read two new datasets, `srtm` and `zion`, using the code below.
    To create a new map representing these datasets.
    
I don't understand what I'm supposed to do here... 

```{r}
srtm = raster(system.file("raster/srtm.tif", package = "spDataLarge"))
zion = read_sf(system.file("vector/zion.gpkg", package = "spDataLarge"))
```


