---
title: "Week 05 - Network Travel and Accessibility"
author: "Adela Sobotkova"
date: "16-February-2022 updated `r Sys.Date()`"
output:
  rmdformats::readthedown:
    highlight: kate
---

```{r setup, include=FALSE}
library(knitr)
library(rmdformats)

## Global options
options(max.print="75")
opts_chunk$set(echo=TRUE,
	             cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```

## Why compute travel times on a network?

* Because you can ....
* Because it matters....

Humans are animals whose activities are constrained by a spatial range. Some of us like to live close to work, some to shopping, others to hotspots of social activity or green areas. While people differ in specific priorities, they all tend to optimize the cost of movement.
In this tutorial, based on Kyle Walker's [MUSA class](http://walker-data.com/MUSAmasterclass/tutorial/#421_Obtaining_demographic_data_with_tidycensus) and mapboxAPI package, we learn to use mapboxAPI to calculate ease of travel across modern road networks so that you see how your environment constrains and impacts your daily behavior. 


### Libraries

To get started with `mapboxapi`, you’ll need to first install some packages. `mapboxapi` was released to CRAN in 2020, so we can install with `install.packages()`. Beware that the process can be a bit entailed and require several updates. That is at least what it took me on a W10 machine running R 4.1.1  The result is worth it, but allow for enough time and restart if R hangs up in the installation/update process. It is also good to install or update the `Rcpp` library to >1.0.7  version. 

```{r packages}
# If running this the first time, uncomment these lines

# install.packages("Rcpp")
# install.packages("fasterize")
# install.packages("mapboxapi", dependencies = TRUE)
```

## MapboxAPI
Before we get started using Mapbox services in R, you’ll need a valid Mapbox account with an access token. Fortunately, Mapbox has generous allowance of 100,000 free API requests so you should not need to pay anything for the API use. To set up your account, visit https://account.mapbox.com/auth/signup/ to establish an account - all you need to provide is an email address to sign up! Fill out the form and verify your account through the email Mapbox sends you; you’ll be taken directly to your Mapbox account dashboard page.

Note the “default public token” that appears on your dashboard screen - you’ll come back to this page in a moment. First, look to the right side of your screen and click “View billing” under Plan. This is where Mapbox will handle your billing information. Nothing you’ll do today will be intensive enough to incur charges - but if you plan to do serious work, you need to be aware that mapboxapi is not a forever free service. Copy the access token that appears on your screen to your clipboard, then return to R.


All features in mapboxapi require a valid Mapbox access token to work. Now that you have yours in hand, you can set yours up! Load the mapboxapi package and install your token as follows: 


```{r mapbox-auth, eval = FALSE}
# either paste it in directly, or in a txt file that you store locally and "ignore"
my_token <- "YOUR TOKEN GOES HERE"
my_token <- read_file("mytoken.txt")

library(mapboxapi)
mb_access_token(my_token, install = TRUE, overwrite = TRUE)
```

The optional argument `install = TRUE` saves the token to your .Renviron, allowing you to use mapboxapi functions in the future without having to worry about setting your token. To use this feature, *restart your R session*.


### Mapbox maps

As you might expect, one of the features of Mapbox services is its ability to create neat web maps.
Why do we care when we can play with leaflet() and tmap() and other libraries? It is important as the Mapbox [Terms of Service](https://www.mapbox.com/legal/tos/) require that Mapbox API outputs be visualized on Mapbox maps and we will be creating MapboxAPI outputs, such as isochrones, accessibility and navigation tools.

Mapbox maps are accessed through styles, which are custom design configurations applied to OpenStreetMap or even user-generated vector map tilesets. You’ll learn how to create and use your own map style with Mapbox later in this workshop. However, Mapbox provides a number of their styles to all users with a Mapbox access token. The most recent versions of these styles (as of the workshop date) are as follows:

* streets-v11: The core Mapbox Streets basemap
* outdoors-v11: A basemap designed for outdoor recreation uses
* light-v10: A light, greyscale background suitable for thematic overlay
* dark-v10: A dark basemap suitable for thematic overlay
* satellite-v9: A global satellite basemap derived from MODIS, Landsat, & proprietary imagery sources
* satellite-streets-v11: The satellite basemap with a streets overlay

You already know one of the most popular R packages for interactive data visualization in R, the Leaflet package maintained by RStudio, which wraps the Leaflet JavaScript library for web mapping.
mapboxapi provides a convenience function, `addMapboxTiles()`, to help you use Mapbox styles in Leaflet in a straightforward way. This function queries the [Mapbox Static Tiles API](https://docs.mapbox.com/api/maps/static-tiles/) and converts a Mapbox style into static tiles for web mapping.


Let’s load the leaflet and mapboxapi libraries and set up an interactive map:
*Please note: your map won’t show up in the RStudio Viewer pane; pop it out to a web browser to view it. It should show in the rmarkdown in its own time*

```{r mabox-map}
library(leaflet)
library(mapboxapi)

mapbox_map <- leaflet() %>%
  addMapboxTiles(style_id = "streets-v11",
                 username = "mapbox") 

mapbox_map

```

### Geocoding with MapboxAPI - Single locations

Once we’ve set up our Leaflet map with a Mapbox basemap, we’ll likely want to focus it on a specific location. mapboxapi includes functionality for R users to interact with the Mapbox Search API. Implemented functions include `mb_geocode()` for *forward geocoding*, which refers to the conversion of a description of a place (like an address) into longitude/latitude coordinates; and `mb_reverse_geocode()`, which converts coordinates into a place description.

Both functions default to using the `mapbox.places` API endpoint, which is to be used for *temporary geocoding*. This means that the endpoint cannot be used to store geocoded information nor can it be used for *batch geocoding* (e.g., a spreadsheet of addresses). These tasks are permissible with the `mapbox.places-permanent endpoint`, which is not included with free accounts. 
R users looking for free batch-geocoding solutions should use other packages like the `tidygeocoder` or `opencage` package. Mapbox geocoding with the `mapbox.places` endpoint *can* be used to focus web maps and guide  navigation services, which will be illustrated in the following sections.

Let’s use `mb_geocode()` to identify the coordinates representing the Aarhus University Nobelparken campus.

```{r geocode}
nobel <- mb_geocode("Nobelparken, Jens Chr.Skous Vej 5, Aarhus, 8000, Denmark ")
nobel
```
By default, `mb_geocode()` returns a length-2 vector representing the longitude and latitude coordinates of the geocoded location. The function can also return an sf POINT object or an R list representing the full API response, if requested. Using the returned coordinates, we can focus our Leaflet Mapbox map with the `setView()` function:
```{r mapbox-nobel}
mapbox_map %>% 
  setView(lng = nobel[1],
          lat = nobel[2], 
          zoom = 14)

```

### Quick Exercise

**Try it out!** Make a Leaflet map in R using a Mapbox basemap of *your* choice, focused on a location of your choice. For locations in non-English-speaking countries: `mb_geocode()` has a `language` argument that can be used to improve the accuracy of queries in languages other than English. [Supported languages (and how to specify them) are found in the Mapbox documentation here](https://docs.mapbox.com/api/search/geocoding/#language-coverage).

```{r}
## A QUICK MAP OF YOURS
```

### Navigation APIs in R

The Mapbox Navigation Service API includes a variety of methods for performing routing and network analysis. mapboxapi allows R users to interact with the Navigation Service API by using the following functions:

* `mb_directions()`: an interface to the Mapbox Directions API for calculating driving (with or without traffic), walking, or cycling routes. Users can request routes between an origin/destination pair or along a series of points, and can return route linestrings as simple features objects along with travel instructions.
* `mb_optimized_route()`: duration-optimized routing with the Mapbox Optimization API, helping you determine the fastest way to visit multiple locations on your route.
* `mb_isochrone()`: uses the Mapbox Isochrone API to draw isochrones around specified locations, which represent the reachable area from those locations within a given travel time by a given travel mode.
* `mb_matrix()`: for a specified set of origin and destination locations, uses the Mapbox Matrix API to calculate pairwise travel times between those locations. This function is recommended for smaller travel-time matrices; large matrix requests (e.g. with results exceeding the tens of thousands) should be completed with a user-installed routing engine like Valhalla or OSRM.


The fifth Mapbox navigation service, map matching, is not yet supported by mapboxapi but will be in a future release.

### Drawing isochrones with Mapbox and R
Creating and visualizing isochrones is beautifully straightforward with the `mb_isochrone()` function in mapboxapi. Supported travel profiles include driving (with no traffic), cycling, and walking. `mb_isochrone()` by default returns a simple features polygon object that can be used for visualization and even spatial analysis.

Let’s try drawing isochrones around the Nobelparken. `mb_isochrone()` accepts an an input a coordinate pair, a location description as a character string, or an sf object. We can use our `nobel` object here to initialize the isochrones around campus.
```{r nobel-drive}
nobel_iso_drive <- mb_isochrone(nobel,
                                 profile = "driving",
                                 time = c(4, 8, 12))

nobel_iso_drive
```
An sf object is returned with a `time` column representing the travel-time around the location. `time` is organized in descending order to ensure that overlapping isochrones are plotted correctly, with the shortest time visualized last (on top).

Using Leaflet’s `addPolygons()` function, we can add the isochrones to our map


```{r nobel-drive-map}
colors <- viridisLite::viridis(3)

# You can set your own colors, too
# library(RColorBrewer)
# colors <- c('#fc8d59','#ffffbf','#91bfdb')


leaflet() %>%
  addMapboxTiles(style_id = "outdoors-v11",
                 username = "mapbox") %>%
  addPolygons(data =nobel_iso_drive, 
              color = rev(colors),
              fillColor = rev(colors),
              fillOpacity = 0.5, 
              opacity = 0.5, 
              weight = 0.2) %>% 
  addLegend(labels = c(4, 8, 12),
            colors = colors,
            title = "Drive-time<br/>around Nobel")
```
### Routing with mapboxapi

mapboxapi can also be used to quickly represent and visualize routes between two locations, or alternatively along multiple locations. At its simplest, however, `mb_directions()` just requires an origin and a destination:

```{r route-moesgaard}
route <- mb_directions(origin = nobel,
                       destination = "Moesgaard Museum, Aarhus, Denmark",
                       profile = "cycling")

mapbox_map %>%
  addPolylines(data = route, 
               popup = paste0(
                 "Distance (km): ",
                 round(route$distance, 1), 
                 "<br/>Time (minutes): ",
                 round(route$duration, 1)
               ))
```

The optional argument `steps = TRUE` will break the route object into separate rows for each leg of the trip, and return travel instructions in a number of different languages (English is the default).

```{r directions-moesgaard}
route_dir <- mb_directions(origin = nobel,
                       destination = "Moesgaard Museum, Aarhus, Denmark",
                       profile = "cycling",
                           steps = TRUE)

route_dir
```

### Exercise 1 - Navigation

Now that you’ve learned how to use isochrone and routing services in mapboxapi, try them out for yourselves! Create the following maps:

1. An isochrone map around a location of your choice. Times can be specified at 1-minute intervals all the way up to 60 minutes using a vector.
2. A route between two locations of your choice, using a travel profile of your choice.

```{r}
## YOUR CODE HERE
```

How many and which Viking monuments are reachable within each of the three bands?
```{r}
## YOUR CODE HERE
```

## Task 1: Analyzing Accessibility - Skejby Hospital 
Resource accessibility is a major issue in modern society. While equality of access is a fundamental tenet of most democracies, the realities of geography, differential biology and income collude to prevent such equitable access. Democracies concerned for their citizens thus need to interrogate service availability to everyone regardless of location, social status and other social markers and compensate for its deficiencies. 

Our class is taking place only a few months after the 2021 communal and regional elections in Denmark, where a number of issues touched on accessibility of medical care. One of the major issues on the ballot was the centralisation of hospitals in the Midtjylland region. The argument for centralisation being the better quality and greater efficiency of services for inpatients, the con being the excessive travel times for citizens residing in more remote parts of Jutland.

Hospital accessibility can be analyzed using Mapbox services and the mapboxapi package in sf-based data science workflows.

In this exercise we can explore one or more advanced applications of mapboxapi. We'll probably not get beyond 2 but let's see :) 

1. We’ll examine how to visualize accessibility to a Skejby hospital in Aarhus; 
2. Identify areas where populations may have difficulty reaching medical services in central Jutland; and 
3. Build a routing app with Shiny that identifies the closest hospital to a user’s address. This section may include some new concepts or techniques - but it is designed to illustrate where you can go with mapboxapi in your work!

### Visualizing the (in)accessibility of Skejby
We can visualize this situation in central Jutland with layered isochrones. We already used this technique to show multiple drive times around the Aarhus University earlier in this tutorial. In this case, we will use `mb_isochrone()` to generate dozens of isochrones, then visualize them simultaneously to illustrate an accessibility gradient in the region.

We’ll first generate the isochrones using a vector of times, 1 through 45 at 1-minute intervals, around the Skejby hospital emergency room address.

```{r isochrones-Skejby, eval =FALSE}
library(mapboxapi)

isos <- mb_isochrone(
  location = " Palle Juul-Jensens Blvd. 161, 8200 Aarhus",
  profile = "driving",
  time = 1:45
)

```
Next, we can visualize our overlapping isochrones. We’ll use the `viridis` color palette as we did previously in the tutorial, and generate a color palette derived from the `time` column in our dataset. Once specified, we can add these polygons to our Mapbox basemap with a mostly-transparent fill opacity.

```{r load-isos, echo=FALSE}
library(sf)
isos <- read_sf("../data/isos4326.shp")
```


```{r map-Skejby-isochrones}
pal <- colorNumeric("viridis", isos$time, na.color = "transparent")

mapbox_map %>%
  addPolygons(data = isos,
              fillColor = ~pal(time),
              stroke = FALSE,
              fillOpacity = 0.1) %>%
  addLegend(values = isos$time,
            pal = pal,
            title = "Drive-time to Skejby hospital")
```

The result illustrates some of the wide differences in accessibility between various parts of the region. One notable issue with this visualization approach, however, is that the layering of isochrones makes it difficult to view the basemap beneath them. This can be resolved by converting to a raster dataset and generating an “accessibility surface” for improved visualization.

### Let's save the isochrone layers
To make continuation easier and save on the API calls.
```{r save-iso}
# dir.create("data")
# library(sf)
# write_sf(isos, "data/isos4326.shp")
```


### Making an “accessibility surface”
*Accessibility surfaces* are commonly used in geographic information systems applications to identify the distance from any particular location to a geographic feature of interest. We can apply this concept to network-based accessibility by using mapboxapi tools. To create the accessibility surface, we will convert our isochrones to a raster dataset using the fasterize package. Raster datasets represent geographic information as grid cells defined by a cell size. Higher-resolution raster datasets are represented with smaller cell sizes.

To generate the accessibility surface raster, we will need to apply a *coordinate system transformation* to “project” our data to two-dimensional coordinates. This will allow us to specify the raster’s resolution in meters. We generate a 100m resolution raster, and use the `fasterize()` function to allocate the minimum overlapping value from our isochrones to each grid cell. The result can then be mapped with Leaflet’s `addRasterImage()` function.

```{r rasterize-isochrones, eval = FALSE}
library(Rcpp)
library(fasterize)
library(raster)
library(sf)

isos_proj <- st_transform(isos, 25832)

template <- raster(isos_proj, resolution = 100)
iso_surface <- fasterize(isos_proj, template, field = "time", fun = "min")

pal <- colorNumeric("viridis", isos$time, na.color = "transparent")

mapbox_map %>%
  addRasterImage(iso_surface, colors = pal, opacity = 0.5) %>%
  addLegend(values = isos$time, pal = pal,
            title = "Drive-time to Skejby Hospital")

```

```{r rasterize-isochrones-sol, echo = FALSE}
library(raster)
iso_surface <- raster("../data/iso_surface100.tif")

pal <- colorNumeric("viridis", isos$time, na.color = "transparent")

mapbox_map %>%
  addRasterImage(iso_surface, colors = pal, opacity = 0.5) %>%
  addLegend(values = isos$time, pal = pal,
            title = "Drive-time to Skejby Hospital")
```

And again, let's save the accessibility surface for later
```{r save-raster}
# library(raster)
# ?writeRaster()
# writeRaster(iso_surface100, "data/iso_surface100.tif", format = "GTiff" )
```

Ok, task one down. Onto task two.

## Task 2: Analyzing Accessibility - Hospitals in Midtjylland
So the accessibility of Skejby looks pretty good, plus it is inside a big city which has a lot of public transport options. What about more rural regions of Midtjylland and their access to medical care?

You can find a list of all hospitals in Denmark in https://www.listsclub.com/hospitals-in-denmark/.
For Midtjylland, the list is: 

* Regionshospitalet Brædstrup in Brædstrup
* Regionshospitalet Grenaa in Grenaa
* Regionshospitalet Hammel Neurocenter in Hammel
* Regionshospitalet Herning in Herning
* Regionshospitalet Holstebro in Holstebro
* Regionshospitalet Horsens in Horsens
* Regionshospitalet Kjellerup in Kjellerup
* Regionshospitalet Lemvig in Lemvig
* Regionshospitalet Odder in Odder
* Regionshospitalet Randers in Randers
* Regionshospitalet Ringkøbing in Ringkøbing
* Regionshospitalet Samsø on the island of Samsø
* Regionshospitalet Silkeborg in Silkeborg
* Regionshospitalet Skanderborg Sundhedscenter in Skanderborg
* Regionshospitalet Skive in Skive
* Regionshospitalet Tarm in Tarm
* Regionshospitalet Viborg in Viborg
* Aarhus Universitetshospital Skejby in Aarhus


Use the opencage library to geocode these. Opencage is an API service, so you will again need to sign up in order to get an API Key for 2500 *free* API requests per day at 1 per second rate. Installation and registration of opencage is much less entailed than mapboxapi and should work straitforwardly.

If short on time, skip to "Plot geocoded data" section and continue with risk assessment.

## Opencage API
You have read about geocoding above in the mapboxapi section, and learnt that mapboxapi does not let you batch-geocode addresses. Opencage does with familiar-sounding `oc_forward_df()` and `oc_backward_df()` which take a single address or a vector of string addresses! 
The next steps will take you through the opencage API package to geocode a list of hospitals in order to get their coordinates.

* Register at [opencagedata.com/users/sign_up](https://opencagedata.com/users/sign_up).
* Generate an API key at the [OpenCage dashboard](https://opencagedata.com/dashboard#api-keys).
* Save your API key as an environment variable like `OPENCAGE_KEY=yourkey` in `.Renviron`. See `help(oc_config)` for alternative ways to set your OpenCage API key.

If you want to know more, the opencage vignettes are really great! There are “Introduction to opencage” [vignette("opencage")](https://docs.ropensci.org/opencage/articles/opencage.html);  “Customise your query” [vignette("customise_query")](https://docs.ropensci.org/opencage/articles/customise_query.html) and “Output options” [vignette("output_options")](https://docs.ropensci.org/opencage/articles/output_options.html)
```{r opencage-auth, eval = FALSE}
library(opencage)
library(tidyverse)

# see configuration options
help(oc_config)

# set key in the environment
Sys.setenv(OPENCAGE_KEY = "YOUR KEY")

# set the key interactively if it is missing.
oc_config()

```

### Geocoding hospitals
Now you are ready to turn place names into latitude and longitude coordinates:
```{r geocode-openc}
b <- oc_forward_df(placename = "Regionshospitalet Grenaa in Grenaa", countrycode = "DK")
leaflet() %>% 
  addTiles() %>% 
  addMarkers(b$oc_lng, b$oc_lat)
```

Let's try with a vector of addresses. You might need to clean the list above in regex101.com before you make it into a character list!  Alternatively, check out [this](https://regex101.com/r/kbGyPh/1)

```{r hospital-list}
hospitals <- c("Regionshospitalet Brædstrup in Brædstrup","Regionshospitalet Grenaa in Grenaa","Regionshospitalet Hammel Neurocenter in Hammel","Regionshospitalet Herning in Herning","Regionshospitalet Holstebro in Holstebro","Regionshospitalet Horsens in Horsens","Regionshospitalet Kjellerup in Kjellerup","Regionshospitalet Lemvig in Lemvig","Regionshospitalet Odder in Odder","Regionshospitalet Randers in Randers","Regionshospitalet Ringkøbing in Ringkøbing","Regionshospitalet Samsø on the island of Samsø","Regionshospitalet Silkeborg in Silkeborg","Regionshospitalet Skanderborg Sundhedscenter in Skanderborg","Regionshospitalet Skive in Skive","Regionshospitalet Tarm in Tarm","Regionshospitalet Viborg in Viborg","Aarhus Universitetshospital Skejby in Aarhus")
```

Batch-geocode!
```{r geocode-h, eval=FALSE}
h <- oc_forward_df(placename = hospitals, countrycode = "DK", limit =1, no_annotations = TRUE)

# Look at the object
h

# Look at the map
leaflet() %>% 
  addTiles() %>% 
  addMarkers(h$oc_lng, h$oc_lat, popup = h$oc_formatted)

# Save object
# write_csv(h, "data/midty_hospitals.csv")
```

### Plot geocoded data
Well, the list is not perfect in all instances. Sometimes it gets the hospital , other times it places a marker in the middle of a town. If you want to fix it, provide street addresses and re-geocode.
```{r h-map, echo = FALSE}
h <- readRDS("../data/hospitals.rds")
leaflet() %>% 
  addTiles() %>% 
  addMarkers(data = h, popup = h$oc_formatted)
```


### Accessibility of Midtjylland hospitals
Let's convert the tibble to an sf POINT object. This dataset can be used to analyze which areas are immediately covered by hospital services, and which are not. We’ll measure accessibility using isochrones as above, and consider a 20 minute walk-time, cycling- and driving-time around each hospital. `mb_isochrone()` can accept sf objects as input, and will retain an ID from the input sf object if the column name is specified.

```{r isochrones-h, eval = FALSE}
# Convert to sf object
h_sf <- st_as_sf(h, coords = c("oc_lng", "oc_lat"), crs = 4326) #%>% 
#  st_transform(crs = 25832)


# Walking range
library(opencage)
walking_isos <- mb_isochrone(
  h_sf,
  profile = "walking",
  time = 20,
  id = "placename"
)

# Cycling range
cycling_isos <- mb_isochrone(
  h_sf,
  profile = "cycling",
  time = 20,
  id = "placename"
)

# Driving range
driving_isos <- mb_isochrone(
  h_sf,
  profile = "driving",
  time = 20,
  id = "placename"
)
```

### Plot the accessibility of Midtjylland hospitals
These results can be visualized on our Mapbox map:

```{r map-isos, eval = FALSE}
# Load data if mapboxapi did not work for you
walking_isos <- readRDS("../data/walking_isos.rds")
cycling_isos <- readRDS("../data/cycling_isos.rds")
driving_isos <- readRDS("../data/driving_isos.rds")



mapbox_map %>%
  ___________

```


Well done! Now you can see the reachable area within a 20-minute walk, 20-minute bike ride and a 20-min drive.  You probably don't want to bike for 20 minutes to reach a hospital if not feeling well. Likewise, car driving is comfortable, but not every place is within 20-mins drive of a hospital and not everybody has a car. Finally, remember that these isochrones do not include traffic updates.



### Questions: 

*1. Which areas are not covered by 20-min drive or 20-min bikeride? How big is their total area?*

*2. How much do you need to increase the driving range so that everyone in Midtjylland can access a hospital (let's assume everyone has a car and just want to inform them of how long it will take)?*
