---
title: "Danish Groundwater Nitrates: Kriging and Ecological Farming"
author: "Adela Sobotkova, adapted from Johan Horsmans and Emil Jessen"
date: "April 2026 updated `r format(Sys.time(), '%B %d, %Y')`"
output:
  rmdformats::readthedown:
    highlight: kate
---

```{r setup, include=FALSE}
library(knitr)
library(rmdformats)

## Global options
options(max.print="75")
opts_chunk$set(echo=TRUE,
               cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```

```{r libraries, include=FALSE}
library(sf)
library(terra)
library(tidyverse)
library(gstat)
library(tmap)
```

# Background

Nitrate leaching from agricultural fertiliser is a well-known threat to Danish groundwater. Does organic (ecological) farming really produce lower nitrate levels than conventional farming?

We will:

1. Spatialise nitrate point measurements from Danish boreholes.
2. Krige a continuous nitrate surface across Denmark.
3. Extract predicted nitrate values inside *organic* vs *conventional* agricultural fields.
4. Test whether the two land-use types differ.

If there are no nitrates.csv in your data/ folder, download it from this [sciencedata link](https://sciencedata.dk/themes/deic_theme_oc7/apps/files_sharing/public.php?t=190154e3aa98e3c9dc0cfe726d355963&) and unzip it into a `data/` folder inside Week11.

---

# Task 1: Spatialise the nitrate measurements (~10 min)

`nitrate.csv` holds borehole measurements with a WKT geometry column and values in mg/L. The coordinates are in EPSG:25832 (ETRS89 / UTM zone 32N) but the CSV does **not** carry that information — you need to assign it.

```{r nitrate-load, eval=FALSE}
# Load the CSV and rename the awkward Danish column names
nitrate <- read_csv("../data/nitrate.csv") %>% as.data.frame()

names(nitrate)[names(nitrate) == "Seneste"]       <- "measurement_date"
names(nitrate)[names(nitrate) == "Seneste mg/l"]  <- "nitrate_concentration"
names(nitrate)[names(nitrate) == "Indtag topdybde"] <- "measurement_depth"

# Quick sanity filter:
#   * drop implausibly high concentrations (> 200 mg/L)
#   * keep only shallow boreholes (0 < depth < 15 m) so we measure surface nitrates
nitrate <- nitrate %>%
  filter(nitrate_concentration < 200,
         measurement_depth > 0,
         measurement_depth < 15)
```

## Convert to `sf` using the WKT column

```{r nitrate-sf, eval=FALSE}
# Turn the data frame into an sf object using the WKT column
nitrate <- st_as_sf(nitrate, wkt = __________)

# The CSV has no CRS metadata — assign EPSG  (which one 25832, 32632, or 4326?)
nitrate <- st_set_crs(nitrate, value = _________)

# Verify
st_crs(nitrate)[1]
```

*Why do we `st_set_crs()` here rather than `st_transform()`?*

## Tidy and plot

```{r nitrate-tidy, eval=FALSE}
# Keep measurements between 2012 and 2021, drop duplicated boreholes
nitrate <- nitrate %>%
  select(measurement_date, nitrate_concentration, measurement_depth, WKT) %>%
  filter(measurement_date >= as.Date("2012-01-01"),
         measurement_date <= as.Date("2021-03-10")) %>%
  distinct(WKT, .keep_all = TRUE)

# Add X/Y columns — we will need them as a trend for kriging
nitrate$Easting  <- st_coordinates(nitrate)[, 1]
nitrate$Northing <- st_coordinates(nitrate)[, 2]

# Quick map
ggplot(nitrate, aes(colour = nitrate_concentration)) +
  geom_sf(size = 0.3) +
  scale_colour_gradientn(colours = c("darkgreen", "yellow", "red")) +
  labs(title = "Nitrate measurements 2012–2021", colour = "mg/L") +
  theme_minimal()
```

---

# Task 2: Organic vs conventional fields (~5 min)

## What are "Økologiske arealer"?

The Danish Agricultural Agency publishes an annual shapefile of all registered **ecological (organic) fields** — `Oekologiske_arealer_<year>.shp`. The full dataset `Markblokke_2022.shp` contains **every** registered field block in Denmark. A field is *conventional* if it appears in `Markblokke` but not in any of the organic layers.

For this exercise we use the **2012-2022 snapshot** as a representative organic layer and derive conventional fields by spatial anti-join. The separate years for 2012–2020 are left as an optional extension at the bottom.

```{r load-fields, eval=FALSE}
# All registered fields in Denmark (2022)
all_fields <- st_read("../data/Markblokke_2022.shp")

# Organic fields collated through 2022 (but single year is enough for the method if you find that easier)
organic_fields <- readRDS("../data/oekologiske_arealer.rds")


# Conventional = all_fields that do NOT intersect any organic polygon
idx_conv <- lengths(st_intersects(all_fields, organic_fields)) == 0
conventional_fields <- all_fields[idx_conv, ]

# How many were removed as organic?
nrow(all_fields) - nrow(conventional_fields)

# Random subsample of organic fields to balance class sizes and cut compute
set.seed(42)
organic_fields <- organic_fields %>%
  slice_sample(n = nrow(conventional_fields))
```

*Why subsample the organic fields? What would happen to the t-test if we kept all ~600k of them?*

---

# Task 3: Variogram and kriging (~15 min)

```{r variogram, eval=FALSE}
# Empirical variogram with a linear X+Y trend (to detrend the N–S gradient)
vgm_emp <- variogram(nitrate_concentration ~ Easting + Northing, data = nitrate)
plot(vgm_emp, main = "Empirical variogram — Danish nitrates")

# Eyeball sill / range / nugget, then fit a Stein-Matern model
v_model <- fit.variogram(
  vgm_emp,
  model = vgm(model = "Ste",
              nugget = 650,
              psill  = 200,
              range  = 40000,
              kappa  = 0.5)
)

plot(vgm_emp, model = v_model, main = "Fitted variogram")
```

*At roughly what distance does spatial correlation die off? Does a 40 km range feel sensible for a country-scale process?*

## Build a prediction grid and krige

```{r krige-grid, eval=FALSE}
# 2 km regular grid across the nitrate bounding box — modern sf workflow
grid_sf <- st_make_grid(nitrate, cellsize = 2000, what = "centers") %>%
  st_as_sf() %>%
  mutate(Easting  = st_coordinates(.)[, 1],
         Northing = st_coordinates(.)[, 2])

# Kriging interpolation
nitrate_grid <- krige(nitrate_concentration ~ Easting + Northing,
                      locations = nitrate,
                      newdata   = grid_sf,
                      model     = v_model)

# Crop to Denmark and map

# First create DK out of your regional (not municipal) polygons
DK <- readRDS("../data/gadm_______________") %>%
  st_as_sf() %>% 
  st_set_crs(4326) %>%
  _________ %>%  # what do you need to do to get the country' border?
  st_transform(25832)

nitrate_dk <- nitrate_grid[DK, ]

tm_shape(nitrate_dk) +
  tm_dots(col = "var1.pred", palette = "-RdYlGn", size = 0.1,
          title = "Predicted nitrate (mg/L)") +
  tm_shape(DK) + tm_borders() +
  tm_layout(main.title = "Kriged nitrate surface — Denmark")
```

---

# Task 4: Do organic fields have lower nitrates? (~10 min)

Attach each kriged grid point to the field polygon it falls inside, then compare group means.

```{r extract-compare, eval=FALSE}
# Spatial join — each grid cell inherits the field polygon it sits within
organic_pred <- st_join(nitrate_dk, organic_fields, join = st_within, left = FALSE) %>%
  mutate(land_type = "organic")

conventional_pred <- st_join(nitrate_dk, conventional_fields, join = st_within, left = FALSE) %>%
  mutate(land_type = "conventional")

merged <- bind_rows(
  st_drop_geometry(organic_pred)[,      c("var1.pred", "land_type")],
  st_drop_geometry(conventional_pred)[, c("var1.pred", "land_type")]
)

# Quick visual and formal test
boxplot(var1.pred ~ land_type, data = merged,
        ylab = "Kriged nitrate (mg/L)", main = "Organic vs conventional fields")

t.test(var1.pred ~ land_type, data = merged)

lm_model <- lm(var1.pred ~ land_type, data = merged)
summary(lm_model)
```

*Interpret the t-test and the regression coefficient. Are organic fields associated with lower predicted nitrate? Is the difference large enough to matter agronomically?*

---

# Optional extension: Union across multiple organic years

*Skip unless you finish early.* The published organic layers 2012 / 2016 / 2018 / 2019 / 2020 capture different farms in different years. Unioning them gives a more complete "ever-organic" mask, but changes the conventional baseline.


```{r multi-year, eval=FALSE}
years  <- c(2012, 2016, 2018, 2019, 2020)
layers <- lapply(years, function(y) {
  st_read(sprintf("data/Oekologiske_arealer_%d.shp", y)) %>%
    na.omit() %>%
    st_geometry()
})

organic_all <- do.call(c, layers) %>% st_as_sf()

# Re-run the spatial anti-join and the kriging extraction above with organic_all
# in place of organic_fields. Does the t-test result strengthen, weaken, or flip?
```
