---
title: "North and South: Syrian crisis, Danish voters and spatial autocorrelation"
author: "Adela Sobotkova"
date: "04 April 2026 updated `r format(Sys.time(), '%B %d, %Y')`"
output:
  rmdformats::readthedown:
    highlight: kate
---

```{r setup, include=FALSE}
library(knitr)
library(rmdformats)

## Global options
options(max.print="75")
opts_chunk$set(echo=TRUE,
               cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```

```{r libraries, include=FALSE}
library(sf)
library(tidyverse)
library(tmap)
library(spdep)
library(mapview)
```

# Background

Does political preference have a spatial structure in Denmark? Are municipalities that vote similarly located near each other, or are voting patterns scattered at random? This session uses Moran's I to test **spatial autocorrelation** in Danish election results before and after the Syrian crisis.

Data: `data/elections_sp.rds` — a pre-joined sf object (CRS EPSG:32632) with municipality geometry and voting fractions for five parties in 2011, 2015, and 2019.

---

# Task 1: Load spatial and election data (~15 min)

Load the municipal boundary and historical election data `elections_sp.rds`, join them, and make a quick map to verify coverage.

```{r load-data, eval=FALSE}
# Load the pre-joined spatial elections dataset
# elections_sp.rds is an sf object (CRS EPSG:32632, already projected)
# Columns: NAME_2, Party, Y2011/Y2015/Y2019 (total votes),
#          sum2011/2015/2019 (electorate), pct_vote2011/2015/2019 (fraction)
elections <- readRDS("../data/________")

# How many rows? How many distinct municipalities?
nrow(elections)
length(unique(elections$NAME_2))

# How many parties?
unique(elections$Party)

# Map the A.Socialdemokratiet fraction in 2015 to get oriented. 
# Which function helps you find specific strings inside filter()?
elections %>%
  filter(_____("__", Party)) %>%    # Party starts with "A" = Socialdemokratiet
  mapview(zcol = "________")        # e.g. pct_vote2015
```

# Task 2: Explore political geography with tmap (~15 min)

Compare how Danske Folkeparti and Socialdemokratiet vote shares are distributed across Denmark. Map a single year (2015) — then change the year column if time allows.

```{r map-parties, eval=FALSE}
library(tmap)

# Map two parties side by side using tm_facets
# Filter to Danske Folkeparti (Party starts with "O") and
# Socialdemokratiet (Party starts with "A")
elections %>%
  filter(_____("^A|^O", Party)) %>%
  tm_shape() +
  tm_facets(by = "________") +
  tm_polygons("________",
              fill.legend = tm_legend(title = "Vote share 2015"),
             position = tm_pos_out("right", "center")))

# What spatial pattern do you see for Danske Folkeparti?
# Is there a north-south or urban-rural gradient?
```

# Task 3: Spatial autocorrelation — queen contiguity (~30 min)

Are neighbouring municipalities more similar in their Danske Folkeparti support than random pairs? Test this with Moran's I.

## Define neighbours

```{r nb-contiguity, eval=FALSE}
# elections_sp.rds is already projected (EPSG:32632)
# Simplify boundaries slightly for faster computation /rendering
mun_sm <- st_simplify(elections %>% filter(______) ,  # filter so you only get 99 shapes 
                      dTolerance = _____)  # put in a number between 100-250 and observe the difference

# Create queen-contiguity neighbour list
nb <- poly2nb(________, queen = TRUE)
nb

# Get municipality centroids for visualisation
mun_centers <- st_coordinates(st_centroid(mun_sm$geometry))

# Plot the neighbour connections
plot(mun_sm$geometry)
plot(nb, mun_centers, col = "red", add = TRUE)
title("Queen contiguity neighbours")
```

## Moran's I test

```{r moran-df, eval=FALSE}
# Subset to Danske Folkeparti
DF <- elections %>% filter(grepl("^O", Party))

# Row-standardised spatial weights
lw <- nb2listw(nb, style = "W", zero.policy = TRUE)

# Moran's I test (theoretical distribution)
moran.test(DF$________,      # pct_vote2015
           lw,
           zero.policy = TRUE)

# Monte Carlo permutation test (more reliable p-value)
moran.mc(DF$________,
         lw,
         nsim = 999,
         zero.policy = TRUE)

# Interpretation: what is the Moran I statistic?
# Does the p-value allow you to reject the null of no spatial autocorrelation?
# Is DF support positively or negatively autocorrelated in space?
```

# Task 4: Alternative neighbourhood definition — 50 km distance band (~20 min)

Does the result depend on how we define neighbours? Test with a 50 km distance band instead of shared borders.

```{r nb-distance, eval=FALSE}
# Get centroids as sf (needed for dnearneigh)
mun_centers_sf <- st_centroid(mun_sm$geometry, of_largest_polygon = TRUE)

# Neighbours within 0–50 km
nb_50 <- dnearneigh(mun_centers_sf, 0, 50000)

# Visualise
plot(mun_sm$geometry)
plot(nb_50, mun_centers_sf, col = "blue", add = TRUE)
title("Neighbours within 50 km")

# Row-standardised weights
lw_50 <- nb2listw(nb_50, style = "W", zero.policy = TRUE)

# Moran's I with distance-band neighbours
moran.test(DF$________,
           lw_50,
           zero.policy = TRUE)

# Does the conclusion change? Why might island municipalities behave differently
# under contiguity vs. distance-band definitions?
```

---

# Optional Diversion: Cartogram

*Skip this section if you are short on time — it is popular but not required for the main learning objectives.*

A **cartogram** distorts region areas so they are proportional to a chosen variable (e.g. total electorate). Regions with more voters appear larger; rural regions with few voters shrink.

```{r cartogram, eval=FALSE}
library(cartogram)

# cartogram_cont() requires a projected CRS
# Filter to Danske Folkeparti 2015
DF <- elections %>%
  filter(grepl("^O", Party)) %>%
  st_transform(crs = 25832)

# Create cartogram scaled to total electorate in 2015
DF_carto <- cartogram_cont(DF, weight = "________", itermax = 15)
# weight = quoted column name, e.g. "sum2015" for total electorate

# Plot the cartogram
plot(DF_carto["________"],
     main = "Cartogram: Danske Folkeparti 2015\n(area scaled to total electorate)")

# Compare: does the spatial pattern look different from the regular map?
# Which municipalities gained area? Which shrank?
```
