---
title: "Exploring Soil Geochemistry "
author: "Adela Sobotkova"
date: "30 March 2021 updated `r format(Sys.time(), '%d %B %Y')`"
output: 
  html_document:
    toc: true
    toc_float: true
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(sf)
```

## Task 1: Kazanlak geochemical survey data
Your job is to explore the fertility of soils in Kazanlak and apply the knowledge to the places of interest - the places of past human occupation. Did people choose to settle near nutrient-rich areas in the past? The Kazanlak geochemistry project collected soil data in a csv called `kaz_geo.csv` , where `OM_muns` column represents the organic matter content on a scale from 0 - 9 with 0 being nutrient-poor and 9 being nutrient-rich. The ancient settlement centroids can be found in `kaz_scatterpoints.shp`.

### Instructions

* Load the geochemical survey data into `kaz_geo` object.
* Look at the names of columns in the data and get a summary of the numerical OM_muns values. 
* Make a histogram of the organic matter.
* Plot a map of the OM_muns data using ggplot.
* Load `Kaz_scatterpoints.shp` shapefile into `kaz_sites` sf object  
* Plot a map combining sampled locations `kaz_geo` and the `kaz_sites` where soil nutrient data is missing.

```{r ca-data, eval=FALSE}
# Load data

# See what OM_muns measurements are at each location

# Get a summary of the organic matter (OM) values

# Look at the distribution

# Plot a map of organic matter

# Plot both the sampled and unsampled locations
```

It's a little tricky to see spatial trends from the ggplot. Let's see how we can make the patterns clearer.


## Task 2: Fitting a trend surface
The  data shows OM broadly increasing from south-west to north-east. Fitting a linear model with the coordinates as covariates will interpolate a flat plane through the values.

### Instructions

* Find or create coordinates columns in the geochemical survey dataset, `kaz_geo`.
* The response, on the left of the `~` sign, is the name of the column we are modeling.
* The explanatory variables are on the right of the `~` sign, separated by a `+` sign, and are the names of the coordinate columns obtained by `st_coordinates()` in the first step.
* Fit the model and see if the model parameters are significant by seeing stars in the coefficients table.

```{r fit-model, eval=FALSE}
# Are there coordinates and what are they called?

# Complete the formula

# Check the coefficients

```
There seems to be a significant trend! Linear regressions are a good first model for exploring many datasets.

## Task 3: Predicting from a trend surface
Your next task is to compute the soil nutrient content at the ancient village locations. You can use the `predict()` function on the fitted model `m_trend` from the previous chunk for this.

### Instructions

* Create a dataframe with missing data
  - Start with `kaz_sites` and ensure the dataframe has a similar structure and projection to `kaz_geo`.
  - Project and write coordinates to `lat`, `long` columns, create `OM` column and populate it with NAs, and relabel TRAP_Code to `ID`.
  - Assign the result to `kaz_geo_miss`.
* Next, use the  `predict()` function to return predictions at all the original locations.
  - Pass the model `m_trend` as the first argument, as usual.
  - Pass `kaz_geo_miss` to the `newdata` argument to predict missing values.
  - Assign the result to `predictions`.
* Fertile soils are those with higher OM. Our linear model gives us estimates and standard deviation based on a normal (Gaussian) assumption. Compute the probability of the soil nutrient content being over 5 using `pnorm()` with the mean and standard deviation values from the prediction data.

```{r predict, eval=FALSE}
# Create a data frame of missing data


# Predict OM for the missing data
predictions <- predict(___, newdata = ___, se.fit = TRUE)

# Compute the exceedance probability
pFertile <- 1 - pnorm(___, mean = predictions$fit, sd = predictions$se.fit)
```


Filling in the gaps and calculating exceeding values is an important task in many industries.

Plot the result!

```{r}

```

Beautiful! Now you know what the soil nutrient values at your ancient site centroids estimated through a linear trend following a gaussian distribution. But what if you want to take into account spatial weighting?  or, wanted to have not only a point but an entire field of values in the radius of ancient village centroids?

## Task 4: Variogram estimation
You can use the `gstat` package to plot variogram clouds and the variograms from data. Recall:

* The *variogram cloud* shows the differences of the measurements against distance for all pairs of data points.
* The *binned variogram* divides the cloud into distance bins and computes the average difference within each bin.

The y-range of the binned variogram is always much smaller than the variogram cloud because the cloud includes the full range of values that go into computing the mean for the binned variogram.

### Instructions

* You should have the geochem survey data, `kaz_geo` in the memory.
* The gstat `variogram()` function uses the `cloud` argument to plot a variogram cloud. 
* The data is in meters, so use a 10km cutoff to prevent the cloud from being too dense.
* Plot a binned variogram of the non-missing data - the default `cloud` parameter is `FALSE`.

```{r binned-vgm, eval = FALSE}
# Load gstat and sp library
library(_____)
library(_____)

# Reload kaz_geo and miss 

# Make a cloud from the non-missing data up to 10km
plot(variogram(OM_muns ~ 1, ____, cloud = TRUE, cutoff = ___))

# Make a variogram of the non-missing data
plot(variogram(OM_muns ~ 1, ______))
```


In the binned variogram you can see that measurements from sites that are further away from each other are more different.

## Task 5: Variogram with spatial trend
You might imagine that if soil at a particular point is nutrient-rich, then soil one metre away is likely to be fertile too. But can you say the same thing about soil one kilometre away, or ten kilometres, or one hundred kilometres?

The shape of the previous variogram tells you there is a large-scale trend in the data. You can fit a variogram considering this trend with `gstat`. This variogram should flatten out, indicating there is no more spatial correlation after a certain distance with the trend taken into account.

### Instructions
* Use the organic matter in survey data, `kaz_geo` 
* Set the formula for the variogram so that the `OM_muns` value depends on the coordinates.

```{r variogram-spatial, eval = FALSE}
# Remember what projected coordinates are called?

# The OM_muns depends on the coordinates
OM_vgm <- variogram(____ ~ ____ + ____, kaz_geo)
plot(OM_vgm)
```


The plot first levels off after around 4000m, indicating that there appears to be little spatial correlation beyond that distance.


## Task 6: Variogram model fitting
Next you'll fit a model to your variogram. The `gstat` function `fit.variogram()` does this. You need to give it some initial values as a starting point for the optimization algorithm to fit a better model.

The sill is the the upper limit of the model. That is, the long-range largest value, ignoring any outliers.

### Instructions

* You will be using the `OM_vgm` from the previous exercise.
* Estimate some parameters by eyeballing the plot.
  - The `nugget` is the value of the semivariance at zero distance.
  - The partial sill, `psill` is the difference between the sill and the nugget.
  - Set the `range` to the distance where the model first begins to flattens out.     
* Fit a variogram model by calling `fit.variogram()`.
  - The second argument should take the parameters you estimated, wrapped in a call to `vgm()`.
* Plot the binned variogram.

```{r vgm-fitting, eval = FALSE}
# Look at the OM_vgm plot, again. 
plot(OM_vgm)

# Eyeball the variogram and estimate the initial parameters
nugget <- ___
psill <- ___
range <- ___

# Fit the variogram
v_model <- ___(
  OM_vgm, 
  model = ___(
    model = "Ste",
    nugget = nugget,
    psill = psill,
    range = range,
    kappa = 0.5
  )
)

# Show the fitted variogram on top of the binned variogram
plot(OM_vgm, model = v_model)
print(v_model)

```


Plotting a smooth model prediction helps you ignore measurement errors to more easily see the distance at which spatial correlation no longer occurs. You can also apply this model to interpolating the missing values from spatial trends present in existing data, in the next exercise. 