---
title: "Shelters"
author: "Adela Sobotkova"
date: "10/04/2022"
output: html_document
---

```{r setup, include=FALSE}
library(knitr)
library(rmdformats)

## Global options
options(max.print="75")
opts_chunk$set(echo=TRUE,
	             cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```


## MapboxAPI
Before we get started using Mapbox services in R, you’ll need a valid Mapbox account with an access token. Fortunately, Mapbox has generous allowance of 100,000 free API requests so you should not need to pay anything for the API use. To set up your account, visit https://account.mapbox.com/auth/signup/ to establish an account - all you need to provide is an email address to sign up! Fill out the form and verify your account through the email Mapbox sends you; you’ll be taken directly to your Mapbox account dashboard page.

Note the “default public token” that appears on your dashboard screen - you’ll come back to this page in a moment. First, look to the right side of your screen and click “View billing” under Plan. This is where Mapbox will handle your billing information. Nothing you’ll do today will be intensive enough to incur charges - but if you plan to do serious work, you need to be aware that mapboxapi is not a forever free service. Copy the access token that appears on your screen to your clipboard, then return to R.

All features in mapboxapi require a valid Mapbox access token to work. Now that you have yours in hand, you can set yours up! Load the mapboxapi package and install your token as follows: 


```{r mapbox-auth, eval = FALSE}
my_token <- "WRITE YOUR TOKEN HERE"

library(mapboxapi)
mb_access_token(my_token, install = TRUE)
```

The optional argument `install = TRUE` saves the token to your .Renviron, allowing you to use mapboxapi functions in the future without having to worry about setting your token. To use this feature, *restart your R session*.


### Mapbox maps

Let’s load the leaflet and mapboxapi libraries and set up an interactive map:
*Please note: your map won’t show up in the RStudio Viewer pane; pop it out to a web browser to view it. It should show in the rmarkdown in its own time*

```{r mabox-map}
library(leaflet)
library(mapboxapi)

mapbox_map <- leaflet() %>%
  addMapboxTiles(style_id = "streets-v11",
                 username = "mapbox") 

mapbox_map

```

# Load shelters data

```{r}
library(sf)

shelters <- st_read("../data/sheltersAarhus.geojson")

library(mapview)
mapview(shelters)
saveRDS(shelters, "../data/shelters.rds")
```

