---
title: "Week 04 - Distances"
author: "Adela Sobotkova"
date: "06/01/2021 updated`r Sys.Date()`"
output:
  rmdformats::readthedown:
    highlight: kate
---

```{r setup, include=FALSE}
library(knitr)
library(rmdformats)

## Global options
options(max.print="75")
opts_chunk$set(echo=TRUE,
	             cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```

Measuring distance between feature sets is a component of spatial analysis 101 -- a core skill for any analyst. There are several functions in `base` R as well as in the packages `rgeos` and `geosphere` to compute distances, but the `st_distance()` function from `sf` provides a useful feature-to-feature distance matrix as output and can be used for most distance calculation needs.

In this exercise you'll :

* measure the distance from the Nobelparken to the Viking monuments within the Aarhus municipality and identify the 10 closest ones
* measure average distance between ancient inscriptions and the nearest city to gauge what buffer best corresponds to an urban zone of influence


Let's practice measuring distances between simple vector objects! 

# Task 1: Get spatial data

Computing buffers is a key spatial analysis skill and the resulting buffers have a wide range of uses like, for example, identifying the number of roads within one kilometer of a school or computing the number of hazardous waste sites near sensitive natural areas.

Although, technically you can buffer data with unprojected coordinate reference systems, the buffer distance will be more meaningful with a projected CRS so it is highly recommended that you transform unprojected data to a projected CRS before buffering.

## Instructions

* Load the packages `sf` and `raster`
* Load data for Danish municipalities with `getData()` function.
* Create the object `df` in your workspace.
* Use `st_as_sf()` to convert this data frame to an `sf` object, use 4326 for the `crs` argument because these points are latitude/longitude -- an unprojected CRS.
* Use `st_transform()` to transform the points' CRS to match the `municipalities`  so we can plot them together. You will need the `st_crs()` function to get the CRS of the `municipalities` or use the EPSG 25832.

```{r buffer, eval=FALSE}
# Create dataframe with lat long data
df <- data.frame(place = c("Aarhus Domkirke", 
                           "Christianskirken",
                           "Nobelparken"),
                 longitude=c(10.2102364,10.1949197,10.2033485),
                 latitude=c(56.1574919,56.1772228,56.174346))


# Convert the data frame to an sf object             
df_sf <- ___(df, coords = c("longitude", "latitude"), ___)

# Transform the points to match the municipalities' CRS EPSG 25832
df_crs <- st_transform(df_sf, crs = 25832)

# Load municipalities

```

Well done.

# Task 2: Grab the Viking monuments 
Hopefully you have saved the `viking` object from last week locally so as to avoid regenerating it. It is always a good idea to route around tedious steps (such as loading large objects) by saving smaller interim datasets. 
If not, no worries, you can go back to Task 6 of Week 03 (instructions reproduced below) and repeat the steps to have a sf object of Viking-period monuments in Denmark in ETRS89/UTM32N projection that also contains municipality information.

## Instructions from Week 03

* Download the fortidsminder archive from [Sciencedata](https://sciencedata.dk/shared/ce0f8e62af16dab66b45f13be90d00f8) and unzip. 
* Load the anlaeg_all_25832.shp into a `monuments` sf object. It might take couple seconds, because there are 300k of them.
* Filter the monuments object to those from the Viking-era ("Vikingtid")
* Filter the `viking` object to only those within the Aarhus municipality. You can do this either by 

--    a) filtering on the municipality column (which you will have if you have done the spatial join `st_join()` the `viking` to `municipalities_tf`)
--    b) clipping the `viking` object by the Aarhus municipality geometry form `municipalities_tf` object with `st_intersection()`
* Use `head()` to confirm that the new object has municipality information -- for example, it should now have municipality name (`NAME_2`).


```{r monuments-munic, eval=FALSE}
# Load the monuments into R
monuments <- read_sf(" _________")

# Check the periodisation and type of the monuments
_____(monuments$datering)
_____(monuments$anlaegsbet)

# Filter Viking-era finds
viking <- monuments %>% 
  _____(___________)

# Plot the monuments on top of the municipalities
plot(_________(municipalities_tf))
plot(_________, add = TRUE, pch = 16, col = "red")


# Determine whether viking monuments has class data.frame
class(viking)

# Join the viking monuments with the municipalities
viking_neigh <- ________(viking, municipalities_tf)

# Confirm that viking_neigh has the municipality information
head(viking_neigh)

# Filter viking-time monuments to those only in Aarhus municipality
viking_aarhus <- viking_neigh %>% 
  _____(NAME_2 == _________)

```

Great! The monuments object did not have municipality information until you performed the spatial join with `st_join()`. Now that the municipality information is included you can filter the dataset to only those monuments in the Aarhus municipality.

# Task 3: Nobelparken and the Viking-era monuments in vicinity
*This exercise originally appeared in Week 03:Task 8*

Let's calculate the as-the-crow-flies distance between the Nobelparken and the Viking-era monuments.

## Instructions

* Filter Nobelparken out of `df_sf` POINT feature and project to a planar CRS. 
* Test whether the CRS of `nobel` matches the CRS of the `viking_aarhus` with `st_crs()`. If not, reconcile them!
* Use `st_distance()` to compute the distance between Nobelparken and the Viking monuments in Aarhus municipality.
* Use `head()` to take a quick look at the result.
* Identify the index of the 10 sites with the smallest distance. This may take several steps that include `sort()` and `which()` functions.
* Use `nearest10` to pull out the rows from `viking` of the nearest Viking monument (see the `anlaegsbet` variable).

```{r nearest, eval=FALSE}
# Create the nobel object and transform to planar CRS
nobel <- ___(___)

# Test whether the CRS match
___(viking_aarhus) == ___(nobel)


# Compute the distance between Nobel and viking_aarhus
d <- ___(___, viking_aarhus)


# Compute the distance between the Nobelparken and the viking monuments
d <- ________(___________, nobel)

# Take a quick look at the result
________(d)

# Find the index of the 10 nearest viking monuments


# Identify the viking monuments that are the nearest, what kinds of monuments are they?
viking_aarhus[_________,]$anlaegsbet

# Plot the results
```


Nice work! If you look at the result for the last line of code carefully you'll see that the closest viking monument is Befæstning at 1580m. In this case one of our feature sets was a single feature. You may end up applying this function in situations where there are multiple features in both objects. In this situation `sf` will return a matrix. Distance matrices take a long time to calculate and so sometimes you calculate distance between the nearest two features from two different datasets. That is the case of the ancient Roman cities and inscriptions :)



# Task 4: Ancient inscriptions and the nearest cities

*This exercise originally appeared in Homework for Week 03:Task 7*

In this exercise we wish the learn what is the average and median distance between ancient inscriptions (messages written into stone) and ancient cities. This is of interest, because writing is seen as an urban phenomenon and so the presence of inscriptions can help us find buried cities, as well as look at the urban-rural divisions and dynamics.

The best place to start is to go to your homework script and continue with Task 7, or paste the code lines here.

Assuming that you have the object `cities` and `inscriptions` present, please generate a distance matrix and check the average distance between the inscriptions and nearest cities (both should be points) for all the cities within the convex hull of inscriptions. How far from urban centers are the inscriptions on average. Would a small change to the buffer distance dramatically change the urban:rural ratio ?


*Instructions for mean and median distance*

* Ensure the `cities` object has the same CRS as `inscriptions` object.
* Clip or select only those cities that fall within the convex hull of inscriptions to reduce the number of calculations. 
* Calculate the mean distance between the inscriptions and cities with `st_distance()`. 
  --    Reduce the cities object only to those within the convex hull of inscriptions so as to reduce the calculation. 
  --  Calculate the distance pairwise - between the nearest city and inscription only -  rather than between all inscriptions and all cities. Check this documentation to understand the `st_nearest_feature()` function 
https://r-spatial.github.io/sf/reference/st_nearest_feature.html and reduce the computational intensity of your calculation., 
* Run `summary()` on the product of `st_distance()` to see the mean and median distance


```{r}
## YOUR CODE
```


## Questions: 

*1. What is the average distance between the inscriptions and nearest city within the convex hull?*

*2. What is the problem with calculating distances between vector objects with st_distance()?*

*3. What can you say about the spatial distribution of ancient inscriptions vis-a-vis the cities or Viking monuments in Aarhus municipality? What factors might be impacting the locations?*



