---
title: "Week 07 - Map Classification"
author: "Adela Sobotkova"
date: "06/01/2021 updated`r Sys.Date()`"
output:
  rmdformats::readthedown:
    highlight: kate
---

```{r setup, include=FALSE}
library(knitr)
library(rmdformats)

## Global options
options(max.print="75")
opts_chunk$set(echo=TRUE,
	             cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```

##  Learn about Classification methods

This exercise demonstrates some possibilities of classifications in R, based on Ch.9 of [Geocomputation](https://geocompr.robinlovelace.net/adv-map.html)

### Required R libraries

* We will use the sf, raster, and tmap packages.
* Additionally, we will use the spData and spDataLarge packages that provide new datasets.

```{r}
library(sf)
library(raster)
library(tmap)
```


### Data sets 

We will the two familiar data sets:  `munic` and `viking` for the entirety of Denmark. 

* Recreate them from .rds files in your data/ folder and convert to sf objects in 25832 crs.
* Calculate the density of viking monuments per municipality by either
-- getting municipality names into the viking dataset, grouping and tallying the viking points by municipality and then joining to municipalities OR
-- doing a spatial join between the two datasets and then grouping and tallying by name  

```{r munic-viking}

# Load datasets 
munic <- readRDS("../data/gadm36_DNK_2_sp.rds")
viking <- readRDS("../data/viking.rds")

munic <- munic %>% 
  st_as_sf() %>%   # 4326
  st_transform(crs = 25832)  # making it compatible with viking

viking # 25832 CRS

# checking the crs compatibility
all.equal(crs(munic), crs(viking))

viking

# Approach 1: spatial join
munic_viking <- munic %>% 
  st_join(viking)

library(tidyverse)
library(mapview)

munic_viking <- munic_viking %>% 
  group_by(NAME_2) %>% 
  tally()

  #arrange(desc(n))
mapview(munic_viking, zcol = "n")
```


### Existing code

Here are some examples of plotting number of viking monuments in Denmark * Your role is to create a map based on the suggestions below, selecting the most meaningful classification style.

```{r maps}
# This line of code applies the 'pretty' style rounding legend numbers. Try different numbers of classes.

munic_viking <- munic_viking %>% 
  group_by(NAME_2) %>% 
  tally() 


?st_simplify()

plot(munic_viking10000$geometry)

a <- tm_shape(munic_viking) + tm_polygons(col = "n", style = "pretty", n = 5)
a
# "Jenks" style further smooths over the gaps
b <- tm_shape(munic_viking) + tm_polygons(col = "n", style = "jenks", n = 6)
b
# quantile style divides into 5 even groups
c <- tm_shape(munic_viking) + tm_polygons(col = "n", style = "quantile", n=5)
c
# Equal interval style divides the distribution into even groups
d <- tm_shape(munic_viking) + tm_polygons(col = "n", style = "equal", n = 5)
d
# Write maps above to objects and plot them side by side 
# with tmap_arrange() for better comparison

tmap_arrange( a,b,c,d)
```
```{r}
munic_viking10000 <- munic_viking %>%
  st_simplify(dTolerance = 10000)

mapview(munic_viking100)
mapview(munic_viking1000)
mapview(munic_viking10000)
```


#### Questions

* 1. What are the advantages and disadvantages of each classification method?
* 2. Choose the best classification and create a map with easily legible legend and all other essentials.(Select a suitable color palette from http://colorbrewer2.org/, north arrow, scale, map title, legend title, reasonable number of breaks in the classification )
* 3. Which method and how many classes did you end up using to display your data? Why did you select that method?
* 4. What principles did you use in arranging the parts of your map layout the way you did and how did you apply these principles?

# Your solution

```{r}
# /Start Code/ #
pdf("municmap.pdf")
tmap_arrange(a,b,c,d)
dev.off()

# /End Code/ #
```

