---
title: "12 - Compute Tree Canopy"
author: "Adela Sobotkova"
date: "07/01/2021 updated `r format(Sys.time(), '%B %d, %Y')`" 
output:
  rmdformats::readthedown:
    highlight: kate
---

```{r setup, include=FALSE}
library(knitr)
library(rmdformats)

## Global options
options(max.print="75")
opts_chunk$set(echo=TRUE,
	             cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```

# Task 1: Compute tree density by neighborhood (I)
In order to compute tree density by neighborhood you need two things. You will need to know the area of the neighborhoods, which you will compute in the next exercise. And you need the **tree counts by neighborhood** which is the focus of this exercise.

You will produce counts of all trees by neighborhood in NYC and create a single data frame with a column for total trees. The result should be a data frame with **no geometry**.

## Instructions

* Load `sf`, `raster`, and `dplyr` in your workspace.
* Load `trees`, `neighborhoods`, `canopy`, and `manhattan` objects from Week 02 data folder
* Use the `count()` function from `dplyr` on the `trees` object to compute total trees by neighborhood. The neighborhood variable is `hood`. Call the new object `tree_counts`.
* Use `head()` to take a quick look at `tree_counts` and see if the geometry variable still exists.
* Remove the geometry from your `tree_counts` data frame.
* The default for the `count()` function is to name the count field n. Use the `rename()` function from dplyr to change the column called `n` to `tree_cnt`.
* Use `hist()` to create a quick histogram of the tree counts.

```{r count-trees, eval=FALSE}
# Load data and libraries
_______

# Read in the trees shapefile
trees <- _____

# Read in the neighborhood shapefile
neighborhoods <- _____

# Read in the tree canopy single-band raster
canopy <- _____

# Read in the manhattan Landsat image multi-band raster
manhattan <- _____

# Compute the counts of all trees by hood
tree_counts <- ___(trees, hood)

# Take a quick look
___(tree_counts)

# Remove the geometry
tree_counts_no_geom <- ____(___, ___)

# Rename the n variable to tree_cnt
tree_counts_renamed <- ___(tree_counts_no_geom, tree_cnt = ___)
  
# Create histograms of the total counts
___(tree_counts_renamed$tree_cnt)

```

Great, you've completed the first step toward computing tree densities. You now have tree counts by neighborhoods and in the next exercise you'll compute the neighborhood areas.

# Task 2: Compute tree density by neighborhood (II)
We have the tree counts (from the previous exercise). In this exercise you will compute neighborhood areas, add them to the neighborhood `sf` object and then you'll join in the non-spatial tree counts data frame from the previous exercise.

## Instructions

* Compute the areas of the neighborhoods with `st_area()` and wrap them in the `unclass()` function to remove the units. Save this object as `areas`.
* Use `mutate()` to add the areas to your neighborhoods object. Call the new variable `area`.
* `left_join()` the `neighborhoods` object (which should now have areas) and the `tree_counts` object that you calculated in the last exercise.
* Use `mutate()` with the given `ifelse()` code to replace `NA`s with 0.
* Create a new column in `neighborhoods` called `tree_density` which is `tree_cnt/area`. 

```{r tree-density, eval=FALSE}
# Compute areas and unclass
areas <- ___(___(neighborhoods))

# Add the areas to the neighborhoods object
neighborhoods_area <- ___(neighborhoods, area = areas)

# Join neighborhoods and counts by shared columns (find the matching one in neighborhoods)
neighborhoods_counts <- ___(neighborhoods_area, 
                            tree_counts_renamed, by = c(___ = "nta")

# Replace NA values with 0 (lookup ifelse() function if needed)
neighborhoods_counts <- ___(neighborhoods_counts, 
                            tree_cnt = ifelse(is.na(tree_cnt), 
                                              0, tree_cnt))

# Compute the density
neighborhoods_counts <- mutate(neighborhoods_counts, 
                               tree_density = ___/___)
```

You're part way there. You have now computed the tree density variable using, in part, the `sf` function `st_area()`. In the next exercises you will compute the average tree canopy values and then compare.

# Task 3: Compute average tree canopy by neighborhood
In the previous exercises you computed tree density by neighborhood using tree counts. In this exercise you will compute average tree canopy by neighborhood as a percentage so that we can compare if the results are similar.

## Instructions

* Use `head()` to confirm that your `neighborhoods` object has the density results from the previous exercises.
* Transform the CRS of the `neighborhoods` object to match the `canopy` object (load the latter if needed).
* Convert the `neighborhoods` object to an `sp` object for use with the `raster` package.
* Compute the mean of `canopy` values by neighborhood using `raster::extract()` with the argument fun = mean (this might take a minute).
* Use `mutate`() to add a new column `avg_canopy` to neighborhoods, which is equal to `canopy_neighborhoods`.

```{r tree-canopy, eval=FALSE}
# Confirm that you have the neighborhood density results
___(neighborhoods_counts)

# Transform the neighborhoods CRS to match the canopy layer
neighborhoods_crs <- ___(neighborhoods_counts, ___ = ___(canopy, ___ = ___))

# Convert neighborhoods object to a Spatial object (optional)
neighborhoods_sp <- ___(neighborhoods_crs, ___)

# Compute the mean of canopy values by neighborhood 
canopy_neighborhoods <- ___(canopy, neighborhoods_sp, fun = ___)

# Add the mean canopy values to neighborhoods
neighborhoods_avg_canopy <- ___(neighborhoods, avg_canopy = ___)
```


Excellent! Note that you transformed the neighborhoods object's CRS. This is actually not strictly necessary because `extract()` can transform CRS on the fly. But it will be needed for plotting and other operations later so doing manually is important here.

# Task 4: Create plots using ggplot2
As an initial review of the data you created you will compute correlations and create histograms and a scatter plot using `ggplot2`.

## Instructions

* Load the `ggplot2` package.
* Create a histogram of total tree density using `ggplot()` with `geom_histogram()`. The object is `neighborhoods_avg_canopy` and the variable is `tree_density`.
* Create a histogram of average canopy (`avg_canopy`) using `ggplot()` with `geom_histogram()`.
* Create a scatter plot of `tree_density` on the x-axis against `avg_canopy` on the y-axis using `ggplot()` and `geom_point()`. Also add a smooth using `stat_smooth()`.
* Compute the correlation between tree density and average canopy with `cor()` by running the given code.

```{r ggplot, eval=FALSE}
# Load the ggplot2 package
___

# Create a histogram of tree density (tree_density)
___(___, aes(x = ___)) + 
  ___(color = "white")

# Create a histogram of average canopy (avg_canopy)
___(___, aes(x = ___)) + 
  ___(color = "white")

# Create a scatter plot of tree_density vs avg_canopy
___(neighborhoods, aes(x = ___, y = ___)) + 
    ___() + 
    ___()

# Compute the correlation between density and canopy
cor(neighborhoods$tree_density, neighborhoods$avg_canopy)
```


Nice! Ggplot2 is great for creating data graphics and in the next exercise you'll see that you can use ggplot2 to make maps. In this case the scatter plot and correlation suggest an unexpected result. If anything, the street tree density data and tree canopy data are negatively correlated. You will confirm this with maps in the next tasks and you will learn why.

# Task 5: Create a map using ggplot2
The `geom_sf()` function operates like any other layer in `ggplot2` where you can link variables to aesthetics on the plot through the `aes()` function. In a mapping context this might mean, for example, creating a choropleth map by color coding the polygons based on a variable. If you leave off the aesthetic mapping `geom_sf()` will map the geometry alone.

`devtools::install_github("tidyverse/ggplot2")`

## Instructions

* Ensure that `ggplot2` is loaded in your workspace and the CRS for both objects match (transform if needed. Note that there is no EPSG code, just a proj4string, for this CRS and this is fine.)
* Use `ggplot()` and `geom_sf()` to create a map of `tree_density` using the default colors.
* Use `ggplot()` and `geom_sf()` to create a map of `avg_canopy` using the default colors.
* Alter the map of `tree_density` by changing the colors with `scale_fill_gradient()`. The hex codes (a `#` followed by numbers) for shades of green have been provided.
* Alter the map of `avg_canopy` by changing the colors with `scale_fill_gradient()`. The hex codes (a `#` followed by numbers) for shades of green have been provided.

```{r map, eval=FALSE}
# Simplify name
neighborhoods <- neighborhoods_avg_canopy

# Plot the tree density with default colors
___(neighborhoods) + 
  ___(aes(fill = ___))

# Plot the tree canopy with default colors
___(neighborhoods) + 
  ___(___(___ = avg_canopy))
  
# Plot the tree density using scale_fill_gradient()
___(neighborhoods) + 
  ___(aes(fill = tree_density)) + 
  ___(low = "#edf8e9", high = "#005a32")

# Plot the tree canopy using the scale_fill_gradient()
___(neighborhoods) + 
  ___(___(___ = ___)) +
  scale_fill_gradient(low = "#edf8e9", high = "#005a32")
```


Great! You're making progress in improving the graphics. The new layer type `geom_sf()` is a big help for creating maps in `ggplot2`. Altering the colors made the maps much more readable and you probably noticed that they seem to show a different pattern. How about doing this with `tmap`? See the next exercise.


# Task 6: Create a map using tmap
The `tmap` package is an excellent tool for making maps. You'll see that it simplifies the process of binning your data values for nice maps and makes it easy to stitch together different layers.

`tmap` operates similarly to `ggplot2` in the sense that it starts with a function, in this case `tm_shape()`, to set up the map and then is followed by layers and other instructions separated with the `+`.

## Instructions

* Load `tmap` library in your workspace.
* Use `tm_shape()` and `tm_polygons()` to create a map of `neighborhoods` with no color-coding.
* Use `tm_shape()` and `tm_polygons()` to create a map of `neighborhoods` and color code by `tree_density`.
* Create a new map by updating your neighborhoods map to include `palette = "Greens"` in `tm_polygons()`.
* Create a similar map of the variable `avg_canopy` to include the following arguments in `tm_polygons()`:
  + `style = "quantile"`
  + `n = 7`.

```{r tmap, eval=FALSE}
# Load tmap library
___(___)

# Create a simple map of neighborhoods
___(neighborhoods) + 
    ___()

# Create a color-coded map of neighborhood tree density
___(___) + 
    ___(___)

# Style the tree density map
___(neighborhoods) + 
    ___("tree_density", palette = ___, 
        style = "quantile", n = 7, 
        title = "Trees per sq. KM")

# Create a similar map of average tree canopy
___(neighborhoods) + 
    ___("avg_canopy", palette = "Greens", 
        style = ___, n = ___, 
        title = "Average tree canopy (%)")
```

You can see the beauty of `tmap`. It makes nice maps in R relatively easily and provides a lot of flexibility to alter your plot as needed

# Task 7: Use tmap to create a final prettier(?) map
In this exercise you will use `tmap` to create a final map with three map views to help you to compare the green you see in an aerial photo with the tree density and average canopy results you computed by neighborhood. The question you're trying to answer is which measure, tree density or average tree canopy, more closely matches what you can see in the aerial photo.

Note: In `map2`, you will make use of the `bbox` argument to force the view extents to match the aerial photo.


## Instructions 1/4

* Remember to load `tmap`

1. Create and view a map of the `manhattan` aerial image with `tm_shape()` and `tm_raster()`.
```{r nice-map-1, eval=FALSE}
# Create a map of the manhattan aerial photo
___(manhattan) + 
  ___()
```

2. Create and view a map of the `neighboorhoods` borders with `tm_shape()` and `tm_borders()`.

```{r nb-borders, eval=FALSE}
# Create a map of the neighborhood polygons
___(neighborhoods) + 
  ___(col = "black", lwd = 0.5, alpha = 0.5)
```


3. Combine the `manhattan` and `neighborhoods` maps above with a `+` and save this as `map1` (the `neighborhoods` layer/map should be second).
4. Create the second map of tree measures with `tm_shape()` and `tm_polygons()`.
5. Combine the two maps with `tmap_arrange()`, use `asp = NA` to map height/width to match the bounding box.

```{r your-code-below, echo = TRUE}
# Your code goes below
```



Congratulations! You've created a nice final map you can share with colleagues to show how average tree canopy is a better representation of green space than the tree density based on the Street Tree Census data from NYC.