---
title: "groundwater_pollution_dk"
author: "Adela Sobotkova adapted from Johan Horsmans and Emil Jessen"
date: "05/06/2021" revised and updated in April 2023
output: github_document
---
# Loading- and preprocessing data:

### Loading packages
```{r}
library(pacman)
p_load(sf, raster, dplyr, tmap, ggplot2, tidyverse, lubridate, sp, gstat, ggthemes, lmtest)
```

### Loading nitrate data:
If you have difficulty running the download.sh script (e.g it is timing out or similar), use this [sciencedata link](https://sciencedata.dk/themes/deic_theme_oc7/apps/files_sharing/public.php?t=190154e3aa98e3c9dc0cfe726d355963&) to download a zipped archive and extract it into a data/ folder inside Week 11
```{r, results='hide'}
# Load the .csv-file as a dataframe and save it as "nitrate":
nitrate <- as.data.frame(read_csv("data/nitrate.csv")) 

# Rename columns to remove spaces and odd characters:
names(nitrate)[names(nitrate) == "Seneste"] <- "measurement_date" 
names(nitrate)[names(nitrate) == "Seneste mg/l"] <- "nitrate_concentration"
names(nitrate)[names(nitrate) == "Indtag topdybde"] <- "measurement_depth" 
```

### Remove outliers (see written report for more info):
```{r}
# Remove all nitrate measures with a nitrate concentration above 200 mg/L:
nitrate <- nitrate %>%
  filter_at(vars("nitrate_concentration"), any_vars(. <200)) # keep those below 200 mg/L

# Remove all nitrate-measures with a measurement depth below 0 (i.e. above ground level) and those below 15 cms ( to get surface values of nitrates):
nitrate <- nitrate %>%
  filter_at(vars("measurement_depth"), any_vars(. > 0)) %>% 
   filter_at(vars("measurement_depth"), any_vars(. < 15))
```


## Processing the nitrate-dataframe to make it compatible with shapefile-format:
```{r}
# Transform coordinate column to characters (redundant as it is a char, but just to make sure):
nitrate$WKT <- as.character(nitrate$WKT)
```

### Option 1:  Making it an sf-object using the WKT column:
```{r}
# Use st_as_sf() with WKT
nitrate <- st_as_sf(nitrate, wkt = "WKT")
```

## Setting CRS (why are we doing that?):
```{r}
# Set the projection of the nitrate data as EPSG 25832:
nitrate <- st_set_crs(nitrate, value = 25832)

# Transform the geometry of the data to the assigned CRS:
# nitrate <- st_transform(nitrate, crs=25832) # not needed as the data was 25832 crs already!

# Verify the data has been assigned new CRS:
st_crs(nitrate)[1]
```

### Filtering dates between 2012 and 2021:
```{r}
nitrate$measurement_date>"2012-01-01"
```

```{r}
nitrate <- nitrate %>%
 select(measurement_date, nitrate_concentration, WKT, measurement_depth) %>%  # substitute geometry for WKT if you went that route
 filter(measurement_date >= as.Date("2012-01-01") & measurement_date <= as.Date("2021-03-10"))
```

### Remove duplicate entries:
```{r}
nitrate <- nitrate[!duplicated(nitrate$WKT), ]
```

### Inspecting "nitrate":
```{r}
head(nitrate)
```

### Plotting points with nitrate per. mg/l metric:
```{r}
ggplot(nitrate, aes(colour = nitrate_concentration)) +
  geom_sf() + 
  coord_sf(datum = st_crs(25832)) + 
  theme_solarized() +
  labs(title = "Nitrate measurements:", color = "Nitrate (mg/L)") + 
  scale_colour_gradientn(colors = c("darkgreen", "yellow", "red"))+
  xlab("Metres") +
  ylab("Metres") #+ theme_bw()
```

### Loading land-use data: ecological and conventional farming fields
```{r}
# Loading "Markblok.shp" which contains all registered fields in Denmark and set CRS as EPSG 25832:
all_fields <- st_read("data/Markblokke_2022.shp") %>% na.omit()

# Convert "fields" to SpatialPolygons-object: WHY?
all_fields <- as_Spatial(all_fields$geometry)

# Load all registered ecological fields (from 2012 to 2020) and set CRS as EPSG 25832::

organic_fields_2012 <- st_read("data/Oekologiske_arealer_2012.shp") %>% na.omit() 
#organic_fields_pre2016 <- st_read("data/Oekologiske_arealer.shp") %>% na.omit() 
#organic_fields_2014 <- st_read("data/Oekologiske_arealer_2014.shp") %>% na.omit() 
#organic_fields_2015 <- st_read("data/Oekologiske_arealer_2015.shp") %>% na.omit() 
organic_fields_2016 <- st_read("data/Oekologiske_arealer_2016.shp") %>% na.omit() 
#organic_fields_2017 <- st_read("data/Oekologiske_arealer_2017.shp") %>% na.omit() 
organic_fields_2018 <- st_read("data/Oekologiske_arealer_2018.shp") %>% na.omit() 
organic_fields_2019 <- st_read("data/Oekologiske_arealer_2019.shp") %>% na.omit() 
organic_fields_2020 <- st_read("data/Oekologiske_arealer_2020.shp") %>% na.omit() 

# Merge organic field-geometries as "organic_fields":
organic_fields <- c(organic_fields_2012$geometry, organic_fields_2016$geometry, organic_fields_2018$geometry, organic_fields_2019$geometry, organic_fields_2020$geometry)

```
### Wrangle the data to get at unique values
```{r}
# Check for spatial duplicates
organic_fields %>%
  st_as_sf() %>%   # distinct() does not work on sfc polygon so we need to convert to sf feature
  dplyr::distinct(x) # there don't seem to be spatial duplicates


## Make a new variable called "conventional_fields", where all polygons in "all_fields" that overlap with the polygons in "organic_fields" are removed:

conventional_fields <- all_fields[lengths(st_intersects(st_as_sf(all_fields),st_as_sf(organic_fields)))==0,]


 # Assess how many organic fields have been filtered out:
length(all_fields) - length(conventional_fields)
```


91 organic fields have been filtered out. In summary, we have now loaded all nitrate-measurements between 2012 and 2021 and all current organic  and conventional fields.

### Inspecting "conventional_fields" and "organic_fields":
```{r}
# Inspect "conventional_fields" and "organic_fields":
head(conventional_fields)

head(organic_fields)
```

### Plotting fields:
```{r}
# Plot organic fields:
plot(organic_fields, main = "Organic fields:")
```

```{r}
# Plot conventional fields:
plot(conventional_fields, main = "Conventional fields:")
```

### Taking subset of conventional fields to reduce computational load (see written report for more info):
```{r}
# Subsetting "conventional_fields" to make it contain as many polygons as "organic_fields":
conv_sample <- st_sample(st_as_sf(conventional_fields), nrow(st_sf(organic_fields)), type = "random")

conventional_fields <- st_join(st_as_sf(conventional_fields), st_sf(conv_sample),
                               join = st_intersects, 
                               left = FALSE)
```

### Plotting nitrate measurement-points on top of organic fields (to assess CRS allignment):
```{r}
plot(st_geometry(nitrate), pch = 16, cex = 0.4, col = "red", main = "Nitrate measurements (red) and organic fields (black):")

plot(organic_fields, add = TRUE)
```

### Plotting points on top of all conventional fields (to assess CRS allignment):
```{r}
plot(st_geometry(nitrate), pch = 16, cex = 0.4, col = "red", main = "Nitrate measurements (red) and conventional fields (black):")
plot(conventional_fields, add = TRUE)
```

### Creating separate columns for nitrate X- and Y coordinates for eventual kriging:
```{r}
nitrate$Easting <- st_coordinates(nitrate)[,1]
nitrate$Northing <- st_coordinates(nitrate)[,2]
```


### Mapview
```{r}
# Mapview to check the lack of overlap
library(mapview)

mapview(st_union(organic_fields)) +
  mapview(nitrate, zcol = "nitrate_concentration")

```


# Kriging:

### Make variogram by fitting X- and Y coordinates to nitrate mg/L:
```{r}
vgm <- variogram(nitrate$nitrate_concentration ~ Easting + Northing, nitrate)
plot(vgm)
```

As seen in the plot above, it appears that it is quite hard to find clear spatial correlations in the data. Nonetheless, it appears that measurements within 55000 meters of each other are more correlated than those further away. As such, we argue that this model can be used for kriging, since it only needs to interpolate data that is close to "real" data" (see written report for more info).

### Fit a model to the variogram:

```{r}
# Use the above variogram to eyeball model parameters:
nugget <- 650 # Initial value.
psill <- 200 # Ceiling - nugget.
range <- 40000 # Point where spatial correlation stops. 

# Fit the variogram to data:
v_model <- fit.variogram(
  vgm, 
  model = vgm(
    model = "Ste",
    nugget = nugget,
    psill = psill,
    range = range,
    kappa = 0.5
  )
)

# Show the fitted variogram on top of the binned variogram (ADELA TEXT)
plot(vgm, model = v_model)
```

### Create kriging grid
```{r}
# Find the dimensions of the nitrate data:
st_bbox(nitrate)

# Using the st_bbox output, manually design a grid slightly larger than the bounding box (for prettier plots). Specify that the kriging resolution should be 2000 meter pr. estimate:
grid <- GridTopology(c(430734,6040448), c(2000, 2000), c(236, 190))


# Make the grid a GridTopology-object with the same CRS as the nitrate data:
gridpoints <- SpatialPoints(grid, proj4string = CRS(projection("+init=epsg:25832 +proj=utm +zone=32 +ellps=GRS80 +towgs84=0,0,0,0,0,0,0 +units=m
+no_defs")))

# Convert gridpoint to SpatialPixels:
spgrid <- SpatialPixels(gridpoints)

# Define coordinate names to be X and Y:
coordnames(spgrid) <- c("Easting", "Northing")

# Plot the grid:
plot(spgrid)

# Use mapview to see the actual extent of the grid in space. Does it overlap Denmark as it should?
mapview(spgrid)
```


### Use fitted variogram model to interpolate new data to the grid defined above


```{r}
# Force the nitrate CRS to be consistent with spgrid (otherwise error messages ensue):
nitrate_sp <- as(nitrate, "Spatial")
crs(nitrate_sp) <- crs(spgrid)

# Do kriging interpolations over the grid
nitrate_grid <- krige(nitrate_concentration ~ Easting + Northing, as(nitrate, "Spatial"), newdata = spgrid, model = v_model)

# Plot the nitrate measurements on top of the kriged "nitrate_grid":

# Plotting raster:
tm_shape(nitrate_grid[1])  +
  tm_raster(title = "Nitrate mg/l", 
            style = "cont",
            palette = "-RdYlGn") +
  tm_credits(text = "Adela") +
  tm_layout(main.title = "Nitrate concentrations in Denmark:") +
  
# Nitrate measurements overlayed on raster:
tm_shape(nitrate) +
  tm_dots()
```

### Load DK map
```{r}
DK <- st_read("data/denmark_administrative_outline_boundary.shp")
```

### Inspect CRS
```{r}
head(DK)
```

We see that the map has the wrong CRS (i.e. 4326).

### Set correct CRS:
```{r}
# Set the projection of the DK-map as EPSG:4326:
DK <- st_set_crs(DK, value = 4326)

# Transform the geometry of the map to the assigned CRS (EPSG:25832):
DK <- st_transform(DK, crs=25832)
```

### Inspect CRS:
```{r}
head(DK)
```
### Crop raster to fit DK map (i.e. remove excess interpolations):
```{r}
nitrate_raster <- raster(nitrate_grid)
nitrate_grid_cropped <- crop(nitrate_raster, extent(DK))
nitrate_grid_cropped <- raster::mask(nitrate_grid_cropped, DK)
plot(nitrate_grid_cropped)

```

### Make plot of interpolated data with DK map and nitrate measurements on top:
```{r}
# Plotting raster:
tm_shape(nitrate_grid_cropped)  +
  tm_raster(title = "Nitrate mg/l", 
            style = "cont",
            palette = "-RdYlGn") +
  
# Adding DK-map:
tm_shape(DK) + 
  tm_polygons(alpha = 0.3) +
  
# Adding nitrate measurements overlayed on raster:
tm_shape(nitrate) +
  tm_dots(col = "grey27") +

# Adding legend, credits and scale bar:
  tm_credits(text = "2023 update") +
  tm_layout(main.title = "Nitrate concentrations in Denmark:",
            legend.position = c("right","top"),
            legend.bg.color = "white", legend.bg.alpha = .2, 
            legend.frame = "gray50",
            bg.color = "lightblue") + tm_add_legend(type = "symbol",  labels = "Nitrate measurements", shape = 20, col = "grey27") + 
  tm_scale_bar(breaks = c(20, 40, 60, 80, 100))
```

# Statistical modeling:

### Extracting all kriged nitrate-values falling within organic fields:
```{r}
# library(raster)
# nitrate_raster <- projectRaster(nitrate_raster, crs = 25832)
# 
# crs(nitrate_raster) <- 25832

organic_fields_prj <- st_transform(organic_fields, crs = crs(nitrate_raster))

# Extract kriged nitrate data (r) that fall "within organic_fields":
organic_extraction <- raster::extract(x=nitrate_raster, y=st_sf(organic_fields), fun=mean, df = TRUE,  na.rm=TRUE)

# Add column land_type and specify that it should be "organic":
organic_extraction$land_type <- "organic"
```


### Extracting all kriged nitrate-values falling within conventional fields:

```{r}
# Extract kriged nitrate data (r) that fall "within conventional_fields":
conventional_extraction <- suppressWarnings(raster::extract(x=nitrate_raster, y=st_as_sf(conventional_fields), fun=mean, df=TRUE, na.rm=TRUE))

# Add column land_type and specify that it should be "conventional":
conventional_extraction$land_type <- "conventional"
```

### Merge the extracted data:

```{r}
# Merge the extracted data as "merged":
merged <- rbind(organic_extraction, conventional_extraction)
```

```{r}

t.test(organic_extraction$var1.pred, conventional_extraction$var1.pred[1:5633])
boxplot(organic_extraction$var1.pred, conventional_extraction$var1.pred)
```

Fit linear regression model with nitrate concentration predicted by land type:
```{r}
# Fit linear regression model with nitrate concentration (var1.pred) ~ land_type:
lm_model <- lm(var1.pred ~ land_type, data = merged)
summary(lm_model)
# Perform coefficient t-test:
coeftest(lm_model, type = "HC1")
```

For interpretation of output, see written report.


