---
title: "Spatial Segregation: Chicago Crime continued"
author: "Adela Sobotkova"
date: "24 March 2022 updated `r format(Sys.time(), '%B %d, %Y')`"
output: 
  html_document:
    toc: true
    toc_float: true
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Task 1: Prepare data for segregation
We can get a more principled measure of the violent crime ratio using a spatial segregation model. The `seg` package implements the theory of spatial segregation (Reardon and O'Sullivan 2004).

`spseg()`is a wrapper function, which calls `spatseg` to compute the set of spatial segregation measures after constructing a population density surface and its local environment parameters with user-specified options. Currently the population density surface is constructed by assuming that the population density is uniform in each census tract.
The returned value from `spseg()` in this case is a list, with `h` and `cv` elements giving the values of the statistic over the input `h` values. The `spatialkernel` package supplies a `plotcv` function to show how the test value varies. The `hcv` element has the value of the best bandwidth. For more information, see [link](https://rdrr.io/cran/seg/man/spseg.html).

Before we segregate, though, let's reduce the dataset to a smaller section.
 
## Instructions I - Create a window to crop the dataset

* Install `seg` package from CRAN, load `spatstat` library. 
* Ensure that `spatstat` is loaded and the `crime_violent` simple feature object is active or read in (check last tutorial). 
* Draw a circular boundary outlining an area of 5000m radius around the list of supplied coordinates for Hilton hotel and make it a simple feature object.
  - You have several options here of drawing a circle. Elaborate on a whiteboard first, as you will likely need all of them later.
  - Inspect visually the result to ensure your circle is inside Chicago city boundary 
* Use the circle simple feature to crop `crime_violent` simple feature object


```{r data-exercise, eval = FALSE}
# Libraries
library(spatstat)

# Data
crime_violent <- ________
Hilton <- __________

# Create a circle with 5000m radius around Hilton
c <- ____________________

# Plot the result to check it fits within Chicago
______
```


## Instructions II - Crop the crime points by new circle

* bring in the old sf object `crime_violent` / `crime_v`
* crop it to the extent of the 5km buffer using the circle you created above
* convert the cropped `crime_centre` result to `ppp` object 
  -   consider using the `disc()` function to create a window, or convert circle c to an `owin` object
  -   use `crimetype` column as a basis for marks, after converting it to factor
* plot the split `crime_violent` ppp object

```{r crop-ppp, eval=FALSE}

# Use the circle to clip crimes
crime_centre <- crime_violent %>% 
  _____________(c) 


# Convert the cropped crimes to ppp object, consider how you can create a window

crime_violent <- ______(st_coordinates(crime_centre), W = ____)
marks(crime_violent) = as.factor(crime_centre$crimetype)
plot(split(crime_violent))
```

# Task 2: Segregation 
Compute the probabilities for violent and non-violent crimes as a smooth surface, as well as the p-values for a point-wise test of segregation. This is done by calling `spseg()` with `smoothing = kernel` .

The `spseg()` function takes data and marks as two separate arguments, and you need to process the marks into a numeric matrix of at least two columns. We shall use `case_when()` again to mutate/create the violent and non-violent columns out of the `crimetype`

## Instructions

* Install the `seg` package
* Ensure the `crime_centre` points data is loaded. 
* Create a column `violent` by assigning crimetype values of "violent" the number 1 and "non-violent" the number 0. Since we need two columns, create a nonviolent column doing the reverse.
  - select the two new columns, drop the geometry and assign the result to `cc` object 
* Run `spseg()` function by
 - assigning the coordinates of `crime_centre` to x,
 - assigning `cc` numeric matrix to data argument and selecting `kernel` as a smoothing argument
* Use `spplot()` to plot the map for the violent/non-violent crime probability in your conference area. 


```{r seg-package-example, eval=FALSE}
#install.packages("seg")
library(seg)

# data needs to be in a numeric matrix
cc <- crime_centre %>% 
  mutate(violent = case_when(
    _________________,
   _________________
    ),nonviolent = case_when(
    _________________,
    _________________
  )) %>% 
  select(violent, nonviolent) %>% st_drop_geometry()

# run the spseg() function
pp <- spseg(x = ______(crime_centre), data=______,
            smoothing = ______, maxdist = 3000)

# Explore result
print(pp, digits = 3)
spplot(pp, main = "Kernel")

# Detach library if you wish to try others
detach("package:seg")
```

Good work! The simulation shows that non-violent crime dominates around my conference location.

