---
title: "Civil Defence Shelters: MAUP and Point Patterns"
author: "Adela Sobotkova"
date: "04 April 2026 updated `r format(Sys.time(), '%B %d, %Y')`"
output:
  rmdformats::readthedown:
    highlight: kate
---

```{r setup, include=FALSE}
library(knitr)
library(rmdformats)

## Global options
options(max.print="75")
opts_chunk$set(echo=TRUE,
               cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```

```{r libraries, include=FALSE}
library(sf)
library(tidyverse)
library(tmap)
library(spatstat)
library(sfarrow)
library(mapview)
```

# Background

Denmark built a network of **176 civil defence shelters** in Aarhus during the Cold War. Are they equitably distributed --- does every neighbourhood have equal access regardless of population density? This session uses shelter locations to explore two foundational spatial methods: the **Modifiable Areal Unit Problem (MAUP)** and **Point Pattern Analysis**.

---

# PART 1: MAUP --- Does the zoning frame change the answer?

# Task 1: Load shelter data

```{r load-shelters, eval=FALSE}
# Load shelter point data (176 civil defence shelters, WGS84)
shelters <- st_read("../data/sheltersAarhus.geojson")

# How many shelters are there?
nrow(shelters)

# Quick interactive overview
mapview(shelters)
```

# Task 2: Create spatial frames and compute equity metric

```{r spatial-frames, eval=FALSE}
# --- Transform shelters to UTM 32N for area-based operations ---
shelters_utm <- st_transform(shelters, crs = 25832)

# --- Frame A: 250 m square grid ---
grid <- st_make_grid(________, cellsize = ________, square = TRUE) %>%
  st_sf(frame = "grid", cell_id = seq_along(.), geometry = .)

# --- Frame B: 500 m hex grid ---
grid_hex <- st_make_grid(________, cellsize = ________, square = FALSE) %>%
  st_sf(frame = "hex", cell_id = seq_along(.), geometry = .)

# --- Frame C: Current urban districts ---
districts <- st_read(________) %>%
  st_transform(crs = 25832) %>%
  rename(cell_id = ________) %>%   # it's good to have an id; check names(districts) to find the right field
  mutate(frame = "historic") %>%
  st_filter(shelters_utm, .join = "intersects")
```

### Count shelters and population per frame/grid cell
```{r count-p-s, eval=FALSE}
# --- Count shelters per unit in each frame ---
# Use lengths(st_contains()) to get a count vector directly
count_shelters <- function(frame_sf) {
  frame_sf %>%
    mutate(n_shelters = lengths(st_contains(frame_sf, shelters_utm)))
}

grid_count <- count_shelters(________)
hex_count  <- count_shelters(________)
hist_count <- count_shelters(________)

# --- Load BBR population data ---
bbr_sf <- sfarrow::st_read_parquet(________) %>%
  st_transform(crs = 25832)

# --- Aggregate est_pop per spatial unit ---
agg_pop <- function(frame_sf) {
  st_join(frame_sf, bbr_sf %>% select(est_pop), join = st_contains) %>%
    group_by(cell_id) %>%
    summarise(pop = sum(est_pop, na.rm = TRUE), .groups = "drop")
}

grid_pop <- agg_pop(________)
hex_pop  <- agg_pop(________)
hist_pop <- agg_pop(________)

mapview(grid_pop, zcol = "pop")
mapview(hex_pop,  zcol = "pop")
mapview(hist_pop, zcol = "pop")

# --- Compute shelters per 100 residents ---
add_equity <- function(count_sf, pop_sf) {
  count_sf %>%
    left_join(st_drop_geometry(pop_sf), by = "cell_id") %>%
    mutate(shelters_per_100 = if_else(pop > 0, n_shelters * 100 / pop, NA_real_))
}

grid_eq <- add_equity(grid_count, grid_pop)
hex_eq  <- add_equity(hex_count,  hex_pop)
hist_eq <- add_equity(hist_count, hist_pop)
```

# Task 3: Map and compare MAUP frames

```{r maup-maps, eval=FALSE}
tmap_mode("view")

make_map <- function(eq_sf, title) {
  tm_shape(eq_sf %>% filter(pop > 0, is.finite(shelters_per_100))) +
    tm_polygons("shelters_per_100",
                style = "pretty",
                fill.legend = tm_legend(title = "Shelters\nper 100"),
                palette = "YlOrRd") +
  tm_shape(shelters_utm) +
    tm_dots(size = 0.5, col = "black") +
    tm_title(title)
}

m_grid <- make_map(________, "250 m grid")
m_hex  <- make_map(________, "500 m hex")
m_hist <- make_map(________, "Urban districts")

tmap_arrange(m_grid, m_hex, m_hist, ncol = 3)

# Interpretation: write 2 sentences below explaining how the apparent
# equity pattern changes across the three frames
```

What other containers would you suggest we use to bin the shelters more equitably?
---

# PART 2: Point Pattern Analysis

# Task 4: Convert shelters to ppp

```{r shelters-ppp, eval=FALSE}
# Project shelters to UTM zone 32N (EPSG:25832) — ppp requires projected CRS
shelters_utm <- st_transform(________, crs = 25832)

# Extract coordinates
coords <- st_coordinates(shelters_utm)

# Define observation window from bounding box
win <- as.owin(st_bbox(shelters_utm))

# Create ppp object (176 points)
shelters_ppp <- ppp(x = coords[, "X"], y = coords[, "Y"], window = win)
shelters_ppp

# Note: edge effects may inflate intensity near the window boundary.
# The K-function envelope in Task 7 uses border correction to address this.
```

# Task 5: Quadrat test

```{r quadrat-test, eval=FALSE}
# Quadrat test: divide window into 5x5 grid, compare observed counts to CSR
qt <- quadrat.test(________, nx = 5)
qt
plot(qt, main = "Quadrat test — Aarhus shelters")

# Interpretation: what does the p-value tell us about whether shelters
# are randomly distributed? Does non-randomness imply equitable coverage?
```

# Task 6: Nearest-neighbour distances and G function

```{r nnd-gest, eval=FALSE}
# Mean nearest-neighbour distance (in metres, UTM projection)
mean_nnd <- mean(nndist(________))
cat("Mean nearest-neighbour distance:", round(mean_nnd), "m\n")

# G function with simulation envelope (99 simulations)
env_g <- envelope(________, Gest, nsim = 99)
plot(env_g, main = "G function — Aarhus shelters")

# Interpretation: does the G curve rise steeply at short distances
# (clustering) or slowly (regularity/CSR)? What does this mean for
# equitable access?
```

# Task 7: K function

```{r kest, eval=FALSE}
# K function with simulation envelope (border correction, 99 simulations)
env_k <- envelope(________, Kest, nsim = 99)

# Linearised plot: subtract Poisson expectation (stored in 'theo' column)
plot(env_k, . - theo ~ r,
     main = "Linearised K function — Aarhus shelters",
     ylab = "K(r) - pi*r^2")

# Interpretation: above zero = more clustering than CSR at that scale.
# What scale(s) show the strongest clustering?
# How does this relate to the equity question from Part 1?
```

---

# Challenge Extension (Optional): Bivariate Point Patterns --- Chicago Crime

*This section is self-contained and independent of Parts 1--2. Skip it if time is short.*

Are arson fires and liquor law violations spatially attracted to each other in Chicago?

```{r chicago-challenge, eval=FALSE}
# Load raw Chicago crime data (268,505 rows)
crimes <- read_csv(________)

# Filter to two crime types
# Note: column is "Primary Type" (mixed case) — backtick quoting required
arson_pts  <- crimes %>% filter(`Primary Type` == "ARSON")
liquor_pts <- crimes %>% filter(`Primary Type` == "LIQUOR LAW VIOLATION")

# Sanity check — these exact counts are expected
stopifnot(nrow(arson_pts)  == 444)
stopifnot(nrow(liquor_pts) == 191)

# Drop rows with missing coordinates
arson_pts  <- arson_pts  %>% filter(!is.na(Longitude), !is.na(Latitude))
liquor_pts <- liquor_pts %>% filter(!is.na(Longitude), !is.na(Latitude))

# Define a common observation window from the combined extent
combined_lon <- c(arson_pts$Longitude, liquor_pts$Longitude)
combined_lat <- c(arson_pts$Latitude,  liquor_pts$Latitude)
win_ch <- owin(xrange = range(combined_lon), yrange = range(combined_lat))

# Create individual ppp objects
pp_arson  <- ppp(arson_pts$Longitude,  arson_pts$Latitude,  window = win_ch)
pp_liquor <- ppp(liquor_pts$Longitude, liquor_pts$Latitude, window = win_ch)

# Superimpose into a single marked ppp (required for Kcross)
pp_combined <- superimpose(arson = pp_arson, liquor = pp_liquor)

# Bivariate K: are arson events attracted to liquor violations?
env_cross <- envelope(pp_combined, Kcross,
                      i = "arson", j = "liquor",
                      nsim = 39,
                      fix.n = TRUE)
plot(env_cross,
     main = "Kcross: Arson events around Liquor law violations")

# Interpretation: if the cross-K curve sits ABOVE the simulation envelope,
# arson events are attracted to liquor violation locations at that scale.
# If below, they repel. If within the envelope, the two crime types are
# spatially independent.
```
