---
title: "01 - Reading in the spatial data"
author: "Adela Sobotkova"
date: "04/01/2021 updated `r Sys.Date()`"
output:
  rmdformats::readthedown:
    highlight: kate
---
# Introduction into GIS in R
There are a few key spatial packages available for Spatial Analysis in R, which you need to install as you progress through these exercises. The most basic are 

* `sf` for working with vector data
* `raster` for working with raster data
* `rgdal` for an R-friendly GDAL interface

The installation of the geospatial libraries GDAL, GEOS and PROJ.4 varies significantly based on operating system (and can be entailed). All of these are dependencies for `sf`, the R package that we will be using for spatial data operations throughout this course. Please follow the Option A: Local installation instructions for your operating system on [this page](https://datacarpentry.org/geospatial-workshop/setup.html) under the heading GDAL, GEOS, and PROJ.4 and continue through the R Packages. Click on the linked pages to make your life easier.

# Task 1: Reading vector data
The `sf` package, created by Edzer Pebesma and colleagues, has dramatically simplified reading vector spatial data into R.

In this exercise you will read in three shapefiles (one point file and two polygon files) using `st_read()`. If you've read in the files correctly, you will see a standard R data frame except it will show some header metadata about the file and you'll see a special `geometry` column which we will discuss later.

## Instructions

* Load the `sf` package.
* All your datasets reside in the 'data' folder.
* Import the `trees` shapefile ("trees.shp").
* Import the `neighborhoods` shapefile ("neighborhoods.shp").
* Import the `parks` shapefile ("parks.shp").
* Use the `head()` function and identify the first few tree species.

## Questions: 

1. How many features do `trees` contain and what kind of geometry are they?
2. How many features are there inside `neighborhoods` and what is their geometry?
3. What is the CRS value in these objects?


## Solution to task 1
```{r load data, echo=TRUE}
# Load the sf package
library(sf)
library(raster)

# Read in the trees shapefile
trees <- st_read("data/trees.shp")

# Read in the neighborhood shapefile
neighborhoods <- st_read("data/neighborhoods.shp")

# Read in the parks shapefile
parks <- st_read("data/parks.shp")

# View the first few trees
head(trees)

# Answers
trees
neighborhoods
crs(trees)
crs(neighborhoods)
```

Well done, you can see how easy it can be to read in shapefiles and you got your first taste of what an sf object looks like.

<br>

# Task 2: Reading raster data
The term "raster" refers to gridded data that can include satellite imagery, aerial photographs (like orthophotos) and other types. In R, raster data can be handled using the `raster` package created by Robert J. Hijmans.

When working with raster data, one of the most important things to keep in mind is that the raw data can be what is known as "single-band" or "multi-band" and these are handled a little differently in R. Single-band rasters are the simplest, these have a single layer of raster values -- a classic example would be an elevation raster where each cell value represents the elevation at that location.

Multi-band rasters will have more than one layer. An example is a color aerial photo in which there would be one band each representing red, green or blue light.

## Instructions

* Load the `raster` package.
* Read in the tree canopy raster with the `raster()` function (this is a single-band raster called "canopy.tif").
* Read in the Manhattan Landsat image with `brick()` (this is a multi-band raster called "manhattan.tif").
* Use the `class()` function to determine the class of each raster object you read in.
* Use the `nlayers()` function to determine how many bands/layers are in each object.

Questions:

1. What are the dimensions (number of columns and rows) of the `canopy` raster?
2. What is the resolution of `manhattan`?



Now you've learned how to read in single and multi-band rasters. You should have noticed, based on the `nlayers()` function, that the `canopy` object has a single layer and the `manhattan` object has three.


## Solution to task 2
```{r loadrasters, echo=TRUE}
# Load the raster package
library(raster)

# Read in the tree canopy single-band raster
canopy <- raster("data/canopy.tif")

# Read in the manhattan Landsat image multi-band raster
manhattan <- brick("data/manhattan.tif")

# Get the class for the new objects
class(canopy)
class(manhattan)

# Identify how many layers each object has
nlayers(canopy)
nlayers(manhattan)

# Answers
dim(canopy)
res(manhattan)
```

