---
title: "05 - Slicing and Dicing Spatial Data"
author: "Adela Sobotkova"
date: "06/01/2021 updated`r Sys.Date()`"
output:
  rmdformats::readthedown:
    highlight: kate
---

```{r setup, include=FALSE}
library(knitr)
library(rmdformats)

## Global options
options(max.print="75")
opts_chunk$set(echo=TRUE,
	             cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```

# Task 1: Dropping geometry from a data frame
One of the great innovations of `sf` over `sp` is the use of data frames for storing spatial objects. This allows you to slice and dice your spatial data in the same way you do for non-spatial data. This means you can, for example, apply `dplyr` verbs directly to your `sf` object.

One important difference between `dplyr` with and without spatial data is that the resulting data frames will include the `geometry` variable unless you explicitly drop it. If you want to force the geometry to be dropped you would use the `sf` function `st_set_geometry()` and you would set the geometry to `NULL`.

## Instructions


* Load the packages `sf` and `dplyr`, and the object `trees`  in your workspace.
* Use the `dplyr` function `count()` to tally the number of trees by species from the `trees` dataset. Sort in decreasing order of `n` by setting `sort` to `TRUE`, and call this new data frame `species_counts`.
* Use the `head()` function to check if your new data frame has a geometry column.
* Drop the geometry column from `species_counts` using `st_set_geometry()`.
* Confirm the geometry column has been dropped by re-running `head()`.

## Solution to task 1
```{r libraries and data,echo=TRUE}
# Libraries and data
library(sf)
library(raster)
library(dplyr)

# Read in the trees shapefile
trees <- st_read("data/trees.shp")

# Read in the neighborhood shapefile
neighborhoods <- st_read("data/neighborhoods.shp")

# Read in the tree canopy single-band raster
canopy <- raster("data/canopy.tif")

# Read in the manhattan Landsat image multi-band raster
manhattan <- brick("data/manhattan.tif")
```


```{r drop-sol, echo=TRUE}
# Create a data frame of counts by species
species_counts <- count(trees, species, sort = TRUE)

# Use head to see if the geometry column is in the data frame
head(species_counts)

# Drop the geometry column
species_no_geometry <- st_set_geometry(species_counts, NULL)

# Confirm the geometry column has been dropped
head(species_no_geometry)
```

Nice work! The `sf` package conveniently allows you to apply `dplyr` verbs to spatial objects. The default of keeping a geometry column is not always what you want, but you can drop that column with `st_set_geometry(data, NULL)`.

# Task 2: Join spatial and non-spatial data
In this exercise you will test joining spatial and non-spatial data. In particular, the `trees` data you have been working with has a full county name (the variable is called `boroname`) but does not have the county codes. The neighborhoods file has both a county name (the variable is called `boro_name`) and the county codes -- neighborhoods are nested within counties. In this exercise, you will create a non-spatial data frame of county name and county code from the neighborhoods object. Then you will join this data frame into the spatial trees object with `inner_join()`.

## Instructions

* Ensure the packages `sf` and `dplyr` and the objects `neighborhoods` and `trees` are loaded in your workspace.
* Use the `select()` function from `dplyr` to limit the `neighborhoods` dataset to the fields `boro_name`, `county_fip` and `boro_code` and save this new object as `boro`.
* Drop the geometry column from `boro` with `st_set_geometry()` and assign it to `boro_no_geometry`.
* Use the `dplyr` function `distinct()` to limit the dataset to just one occurrence of each county/boro (otherwise the county information is repeated for each neighborhood).
* Join the newly created `boro_distinct` data frame into the trees object with `inner_join()` from dplyr.
* Use `head()` to confirm that the new fields (`county_fip` and `boro_code`) are included in the trees object.


## Solution to task 2
```{r innerjoin, echo=TRUE}
# Limit to the fields boro_name, county_fip and boro_code
boro <- select(neighborhoods, boro_name, county_fip, boro_code)

# Drop the geometry column
boro_no_geometry <- st_set_geometry(boro, NULL)

# Limit to distinct records
boro_distinct <- distinct(boro_no_geometry)

# Join the county detail into the trees object
trees_with_county <- inner_join(trees, boro_distinct, by = c("boroname" = "boro_name"))

# Confirm the new fields county_fip and boro_code exist
head(trees_with_county)
```

Excellent! The `sf` package allows you to use common joining and merging functions to combine spatial and non-spatial data. In this exercise you used `dplyr` functions but you could also use a base R function like `merge()`.


# Task 3: Simplify the neighborhood boundaries
In `sf` you can use the `st_simplify()` function to reduce line and polygon complexity. In this exercise you will measure the size of objects before and after `st_simplify()` in two ways. You will compute the size in megabytes using the handy `object_size()` function in the `pryr` package and you will count the number of vertices -- the number of points required to delineate a line or polygon.

## Instructions

* Ensure the packages `sf` and `pryr` are loaded (install them if needed) in your workspace.
* Review the geometry of the `neighborhoods` object.
* Compute the object size for the `neighborhoods` object with `object_size()`.
* Compute the total number of vertices in the `neighborhoods` object by altering and applying the code in the exercise introduction -- `st_cast()`, `sapply()`, `sum()`.
* Use the `st_simplify()` function on neighborhoods and save the new object as `neighborhood_simple`. Use the `preserveTopology = TRUE` argument so that borders stay aligned and use `dTolerance = 100` to set the amount of simplification allowed (this is in meters).
* Compute the object size of `neighborhoods_simple`.
* Compute the number of vertices in `neighborhoods_simple`.
* Review the geometry of the `neighborhoods_simple` object.

## Solution to task 3
```{r simplify-sol, echo=TRUE}
# Library
library(pryr)

# Plot the neighborhoods geometry
plot(st_geometry(neighborhoods), col = "grey")

# Measure the size of the neighborhoods object
object_size(neighborhoods) # 1.5 Mb

# Compute the number of vertices in the neighborhoods object
pts_neighborhoods <- st_cast(neighborhoods$geometry, "MULTIPOINT")
cnt_neighborhoods <- sapply(pts_neighborhoods, length)
sum(cnt_neighborhoods) # 210736

# Simplify the neighborhoods object
neighborhoods_simple <- st_simplify(neighborhoods, 
                            preserveTopology = TRUE, 
                            dTolerance = 100)

# Measure the size of the neighborhoods_simple object
object_size(neighborhoods_simple) # 230 kB

# Compute the number of vertices in the neighborhoods_simple object
pts_neighborhoods_simple <- st_cast(neighborhoods_simple$geometry, "MULTIPOINT")
cnt_neighborhoods_simple <- sapply(pts_neighborhoods_simple, length)
sum(cnt_neighborhoods_simple) #8610

# Plot the neighborhoods_simple object geometry
plot(st_geometry(neighborhoods_simple), col = "grey")
```

Success! Hopefully, this amazed you. The `neighborhoods` object was reduced from nearly 2 megabytes and more than 200,000 points to 215 Kb (0.215 megabytes) and 8614 points but you can barely see the difference in the plots. The size reduction is 8x smaller and 20x fewer points. This is a good way to reduce object size for computation or, for example, for display on the web.


# Task 4: Converting sf objects to sp objects
In order to convert an `sf` object to an `sp` object (which has a Spatial class) you can use the `as()` function with `Class = "Spatial"`. To convert back to `sf` you can use `st_as_sf()` and accept the defaults. This is good to know especially if you plan on doing spatial statistics, as geostatistical packages operate with `sp` objects.

## Instructions

* Read in the `trees` shapefile ("trees.shp").
* Convert the `sf` object `trees` to an `sp` object (class `Spatial`) with `as()`.
* Use the `class()` function to confirm the conversion.
* Convert your new `Spatial` object back to `sf` with `st_as_sf()`
* Use the `class()` function to confirm the conversion.


## Solution to task 4
```{r sf-to-sp-sol, echo=TRUE}
# Read in the trees data
trees <- st_read("data/trees.shp")

# Convert to Spatial class
trees_sp <- as(trees, Class ="Spatial")

# Confirm conversion, should be "SpatialPointsDataFrame"
class(trees_sp)

# Convert back to sf
trees_sf <-st_as_sf(trees_sp)

# Confirm conversion
class(trees_sf)
```
Success! For now you may need to use `as()` and `st_as_sf()` somewhat frequently but as `sp` becomes less common you will be able to stick with `sf` objects.


# Task 5: Converting to and from coordinates
In order to convert a data frame of coordinates into an `sf` object you can make use of the `st_as_sf()` function you used in the previous exercise. You can specify the `coords` argument with the names of the coordinate variables (**with the X coordinate/longitude coordinate listed first**) and, optionally, the `crs` argument if you know the CRS of your coordinates. The CRS can be specified as a proj4 string or EPSG code.

If you want to convert your `sf` point objects to a data frame with coordinates, you can use the `st_write()` function with a hidden argument (these are arguments associated with an external utility called GDAL and so they're not in the R help) to force `sf` to include the coordinates in the output file. The argument you need is `layer_options = "GEOMETRY=AS_XY"`.

## Instructions

* Read in the "trees.csv" file with `read.csv()` and call this `trees`.
* Convert the `trees` object you just created to an `sf` object with `st_as_sf()`. Set the `crs` to 4326.
* Plot the geometry of the points with `plot()` and `st_geometry()` to confirm conversion to `sf` worked.
* Write this points object out to CSV with the coordinates using `st_write()`, call the file `"new_trees.csv"`.
* Read in the CSV you just created with `read.csv()` and confirm that the coordinates exist with `head()`.


## Solution to task 5
```{r csv-to-sf-sol, echo=TRUE}
# Read in the CSV
trees <- read.csv("data/trees.csv")

# Convert the data frame to an sf object
trees_sf <-st_as_sf(trees, coords = c("longitude","latitude"), crs = 4326)

# Plot the geometry of the points
plot(st_geometry(trees_sf))

# Write the file out with coordinates
st_write(trees_sf, "new_trees.csv",  
         layer_options = "GEOMETRY=AS_XY",  #this may no longer be necessary
         delete_dsn = TRUE)  # this may no longer be necessary, delete if you get errors re:source

# Read in the file you just created and check coordinates
new_trees <- read.csv("new_trees.csv")
head(new_trees)
```
Great, working with coordinates by turning them into `sf` objects or writing out to a data frame is a critical skill in spatial analysis. Note that `st_write()` can also be used to write lines and polygons but it wouldn't make sense to write these out with coordinates in the same format.