---
title: "07 - Bboxes, Buffers and centroids"
author: "Adela Sobotkova"
date: "06/01/2021 updated`r Sys.Date()`"
output:
  rmdformats::readthedown:
    highlight: kate
---

```{r setup, include=FALSE}
library(knitr)
library(rmdformats)

## Global options
options(max.print="75")
opts_chunk$set(echo=TRUE,
	             cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```

# Task 1: Buffer layers

Computing buffers is a key spatial analysis skill and the resulting buffers have a wide range of uses like, for example, identifying the number of roads within one kilometer of a school or computing the number of hazardous waste sites near sensitive natural areas.

Although, technically you can buffer data with unprojected coodinate reference systems, the buffer distance will be more meaningful with a projected CRS so it is highly recommended that you transform unprojected data to a projected CRS before buffering.

## Instructions

* Load the packages `sf` and `raster`along with the objects `df` and `manhattan` in your workspace.
* Use `st_as_sf()` to convert this data frame to an `sf` object, use 4326 for the `crs` argument because these points are latitude/longitude -- an unprojected CRS.
* Use `st_transform()` to transform the points' CRS to match the manhattan image so we can plot them together. You will need the `crs()` function to get the CRS of the manhattan image.
* Buffer your points by 1000 meters (no need to specify "meters", since this is the projection unit, this will be the default). The radius argument is `dist`.
* Plot the manhattan image, and the geometries of buffers and points together. This will require running `plotRGB()`, `plot()`, and `plot()` again, all together.


## Solution to task 1

```{r buffer-sol, echo=TRUE}
# Instructor: Create df and load manhattan
library(sf)
library(raster)
library(dplyr)
manhattan <- brick("data/manhattan.tif")

df <- data.frame(place = c("Empire State Building", "Museum of Natural History"),
                 longitude=c(-73.98566,-73.97398),
                 latitude=c(40.74844,40.78132))
#df$geometry <- list(c(585631.7, 4511327),c(586575.1, 4514988))

# Review df
df

# Convert the data frame to an sf object             
df_sf <- st_as_sf(df, coords = c("longitude", "latitude"), crs =4326)

# Transform the points to match the manhattan CRS
df_crs <- st_transform(df_sf, crs = crs(manhattan, asText = TRUE))

# Buffer the points
df_buf <- st_buffer(df_crs, dist = 1000)

# Plot the manhattan image (it is multi-band)
plotRGB(manhattan)
plot(st_geometry(df_buf), col = "firebrick", add = TRUE)
plot(st_geometry(df_crs), pch = 16, add = TRUE)

```

Congratulations! Buffering is pretty straightforward in `sf`. The hardest part is probably getting your data into R and ensuring that they have the right coordinate reference system.

# Task 2: Compute polygon centroids
Similar to buffering, computing polygon centroids is a bedrock geoprocessing task used to assign values and even to help with labeling maps. The function for this in sf is `st_centroid()`.

Also similar to buffering, centroid calculations should generally be performed on data with a projected coordinate reference system.

## Instructions

* Read in the `neighborhoods` shapefile ("neighborhoods.shp").
* Project/transform the neighborhoods shapefile and use CRS number 32618 as the target CRS, this is the code for the CRS of the `manhattan` image.
* Compute the centroids with `st_centroid()` and store as an object named `centroids`. Only one argument is required.
* Plot the geometry of `neighborhoods_tf` object with the centroids on top. This requires two calls to `plot()` run together. The second will need `add = TRUE`.


## Solution to task 2
```{r centroids-sol, echo=TRUE}
# Read in the neighborhods shapefile
neighborhoods <- st_read("data/neighborhoods.shp")

# Project neighborhoods to match manhattan
neighborhoods_tf <- st_transform(neighborhoods, crs = 32618)

# Compute the neighborhood centroids
centroids <- st_centroid(neighborhoods_tf)

# Plot the neighborhood geometry
plot(st_geometry(neighborhoods_tf), col = "grey", border = "white")
plot(centroids, pch = 16, col = "firebrick", add = TRUE)
```

Complete! You will find that knowing how to compute polygon centroids will serve you well in many contexts

# Task 3: Create a bounding box around vector data
You can compute bounding boxes around vector data using `sf`. These can help you, for example, create polygons to clip layers to a common area for an analysis or identify regions of influence.

In the `sf` package, there is a function for extracting the bounding box coordinates, if that's all you need, this is `st_bbox()`. More likely you'll want to create a new `sf` object (a polygon) from those coordinates and to do this `sf` provides the `st_make_grid()` function.

`st_make_grid()` can be used to make a multi-row and multi-column grid covering your input data but it can also be used to make a grid of just one cell (a bounding box). To do this, you need to specify the number of grid cells as `n = 1`.

## Instructions

* Create `beech` trees object by reading in the shapefile ("beech.shp").
* Plot the `beech` trees object on top of the `neighborhoods` object -- wrap `neighborhoods` in `st_geometry()` so you're not plotting any attributes. This requires two calls to `plot()` and one will need `add = TRUE`.
* Compute the bounding box coordinates with `st_bbox()`.
* Create a single polygon bounding box around the beech trees with `st_make_grid()` with an argument of `n = 1`.
* Create a plot of the `neighborhoods` (just the geometry), the `beech` trees and the new box around the beech trees.


## Solution to task 3

```{r beech-create, include=TRUE}
# Instructor to create beech trees
# beech <- trees %>% 
#   filter(grepl('*beech*', species))
# plot(st_geometry(beech))
# st_write(beech, "data/beech.shp")

```


```{r NY-box-sol, echo = TRUE}
# Read in the beech trees shapefile
beech <- st_read("data/beech.shp")

# Plot the neighborhoods and beech trees
plot(st_geometry(neighborhoods), col = "grey", border = "white")
plot(beech, add = TRUE, pch = 16, col = "forestgreen")

# Compute the coordinates of the bounding box
st_bbox(beech)

# Create a bounding box polygon
beech_box <- st_make_grid(beech, n = 1)

# Plot the neighborhoods, add the beech trees and add the new box
plot(st_geometry(neighborhoods), col = "grey", border = "white")
plot(beech, add = TRUE, pch = 16, col = "forestgreen")
plot(beech_box, add = TRUE)
```

Bounding box completed! You can see in the plot that the bounding box is the minimum rectangular polygon that fits all the beech trees. You might now clip a series of layers to this box for a consistent extent. You will learn how to do this in a later exercise.



# Task 4: Dissolve multiple features into one
In order to compute a tighter bounding box, a convex hull, around a set of points like the beech trees from the previous exercise you'll need to learn one more function first.

For points you don't want a convex hull around **each** point! This doesn't even make sense. More likely you want to compute a convex hull around all your points. If you have a set of points and you want to draw a convex hull around them you first need to bundle the points into a single MULTIPOINT feature and in order to do this you will use the dissolve function in `sf` called `st_union()`.

With polygons, `st_union()` will dissolve all the polygons into a single polygon representing the area where all the polygons overlap. Your set of individual points will be dissolved/unioned into a single, MULTIPOINT feature that you can use for tasks like computing the convex hull.

## Instructions

* Buffer the `beech` object by 3000 with `st_buffer()` -- `sf` will automatically use meters, the units of the CRS. But remember to project `beech` to `manhattan` object so as to have units be metric. Use pipes for neater code.
* Create a new object called `beech_buffers` which is just the geometry of the buffered beech trees with `st_geometry()`.
* Compute the number of features in the `beech_buffers` object with `length()` and `plot()` to see what they look like.
* Dissolve the buffers in `beech_buffers`, call this `beech_buf_union`.
* Compute the number of features in the `beech_buf_union` object with `length()` and `plot()` to see what the dissolved object looks like.


## Solution to task 4
```{r buffers-sol, echo=TRUE}
# Buffer the beech trees by 3000; they must be projected to manhattan
beech_buffer <- beech %>% 
  st_transform(crs = crs(manhattan, asText=TRUE)) %>% 
  st_buffer(3000)

# Limit the object to just geometry
beech_buffers <- st_geometry(beech_buffer)

# Compute the number of features in beech_buffer
length(beech_buffers)

# Plot the tree buffers
plot(beech_buffers)

# Dissolve the buffers
beech_buf_union <- st_union(beech_buffers)

# Compute the number of features in beech_buf_union
length(beech_buf_union)

# Plot the dissolved buffers
plot(beech_buf_union)
```

Great! You'll see later that the `st_union()` function is important in other spatial analysis contexts, so it's a good one to commit to memory.

# Task 5: Compute a convex hull around vectors
A more precise bounding polygon is sometimes needed, one that fits your data more neatly. For this, you can use the `st_convex_hull()` function. Note that `st_convex_hull()` will compute a tight box around each one of your features *individually* so if you want to create a convex hull around a group of features you'll need to use `st_union()` to combine individual features into a single multi-feature.

## Instructions

* Use `head()` on `beech` to look at the data frame and see the type of geometry.
* Use `st_union()` to combine the individual points in the `beech` object into a single MULTIPOINT geometry and call this beech1.
* Use `head()` on `beech1` to see the type of geometry of the dissolved object.
* Use the `length()` function from `base` R on `beech` and `beech1` to confirm that the number of features went from 17 to 1.
* Use the `st_convex_hull()` function on `beech1` to compute the tight bounding box around the beech trees and call this `beech_hul`l.
* Plot `beech_hull` and then plot the points on top. Use `plot()` twice and run the lines together.


## Solution to task 5
```{r convexhull-sol, include=TRUE}
# Look at the data frame to see the type of geometry
head(beech)

# Convert the points to a single multi-point
beech1 <- st_union(beech)

# Look at the data frame to see the type of geometry
head(beech1)

# Confirm that we went from 17 features to 1 feature
length(beech)
length(beech1)

# Compute the tight bounding box
beech_hull <- st_convex_hull(beech1)
```


```{r convexhull-sol-plot, echo=TRUE}
# Plot the points together with the hull
plot(beech_hull, col = "red")
plot(beech1, add =TRUE)

```

Hooray! It takes a little getting used to, but when computing a convex hull, you'll need to remember to first dissolve/union/combine your individual features into a multi-feature. When you used `head()` you should have noticed that you started with “POINT” objects and after `st_union()` you had “MULTIPOINT”.

