---
title: "05 - Remote Sensing with Landsat package"
author: "Adela Sobotkova"
date: "06/01/2021 updated`r Sys.Date()`"
output:
  rmdformats::readthedown:
    highlight: kate
---
```{r setup, include=FALSE}
library(knitr)
library(rmdformats)

## Global options
options(max.print="75")
opts_chunk$set(echo=TRUE,
	             cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```

# Goals

In this exercise you learn to use the functions of the landsat package to process satellite imagery, specifically: 

* streamline histograms
* manipulate multi-band imagery,
* extract and create new data from images
* segment features in satellite imagery

# Task 1: Set up your workspace

Start by installing the required packages first. Beware that these are not on worker2, sorry!
RasterVis and RColoBrewer are used for visualisation or raster, lattice and latticeExtra for that of histograms(doublecheck??), while landsat provides the imagery samples and rgl allows for 3D visualisation. 

```{r packages}
install.packages(c("rasterVis", 
                   "RColorBrewer", 
                   "landsat",
                   "lattice",
                   "latticeExtra",
                   "rgl",
                   "itcSegment"))

library(lattice)
library(latticeExtra)
library(RColorBrewer)
library(rasterVis)
library(rgdal)
library(rgl)
```

# Task 2: Pre-processing of Landsat datasets 

Landsat packages offers sample Landsat satellite imagery decomposed into single bands, and labelled by the month of capture (25November2002). You will load and practice raster analysis with these 300x300 pixel samples. 
Note that each single band image shows as panchromatic (greyscale) with values ranging from 0 (white) to 255 (black), with the interim colors being shades of grey. This is unsigned 8 bit imagery. The number behind the filename refers to the color band of the image. 1 = Green, 2 = Blue, 3= Red, 4 = Near-infrared. Once you combine these colorbands and plot with plotRGB(), then you can see the true- or false-color imagery. 

The images are in the SpatialGridDataFrame format and need to be converted to Raster layer before manipulation.


```{r plot-img-data}
library(landsat)

##load indvidual band image data from landsat package
?nov

#load band#3 red channel of the image
data(nov3)
plot(nov3)

data(nov4)
plot(nov4)

```


# Task 3: Load elevation data and plot it in 3D
dem in the landsat package denotes a 'digital elevation model'. This isOnce you load it you can convert it to Formal Raster Layer with raster() function and then continue processing using the usual raster package functions.

plot3D() is a neat function of the rasterVis package, which opens a separate windows and plots elevation values if present in the raster. You need to close the window if you want to update or plot another object.

```{r dem}
# Load and plot the digital elevation model in landsat package
data(dem)
plot(dem)
dem <- raster(dem)

# RasterVis package function plot3D() loads a neat 3D viewer, which opens in a new window
# you may need to enable the rgl widget on a cloud server
plot3D(dem, rev=T, zfac=1)

```


# Task 4: plot RGB data
```{r}
# let's create an RGB image and drape it over the 3D model
data("july1")
data("july2")
data("july3")
data("july4")
j1<-raster(july1) # blue
j2<-raster(july2) # green
j3<-raster(july3) # red
j4<-raster(july4) # near-infrared

## check out the image histogram	
plot(j1)
hist(j1, main="Band 1 - Blue of Landsat")	
boxplot(j1, main="Band 1 - Blue of Landsat")		


### Reorder to R - G - B and create a multi-layer rasterBrick !
myRGB <-	brick(j3,j2,j1) # brick creates new object
myCIR <-	stack(j4,j3,j2) # stack stores connections only


### let's see how the NIR, R, and G bands relate  (from lattice)
splom(myCIR, varname.cex=1) # scatter plot matrix


# let´s plot in full colour!
plot(myRGB)

## better
plotRGB(myRGB)
plotRGB(myCIR, stretch="lin")
```


# Task 5: Manipulate image rendering by histogram stretching
```{r}
## different stretches - here histogram based
plotRGB(myRGB, stretch="hist") 
plotRGB(myCIR, stretch="hist")

## different stretches - here linear stretch 
plotRGB(myRGB, stretch="lin") 
plotRGB(myCIR, stretch="lin")  # in CIR red = green!


###finally...
plot3D(dem, col=rainbow(255)) ## you need to close RGL device manually first and then run this line!
plot3D(dem, drape=j1) ## drapes image j4 over DEM


##### histogram/color adjustments

data(nov3)
data(july3)
par(mfrow=c(2,2))
image(nov3, main="nov")
image(july3,main="july")
plot(nov3, main="nov")
plot(july3, main="july")

nov3.newH <- histmatch(master=july3, tofix=nov3)
image(nov3.newH$newimage, main="new nov")

n3 <- raster(nov3)
j3 <- raster(july3)
n3new <- raster(nov3.newH$newimage)

par(mfrow=c(1,3))

hist(n3, main="Nov"); hist(j3,main="July"); hist(n3new, main="new Nov")

####  most important corrections are atmospheric and topographic
####   however, these are too complex to cover here...see package help

par(mfrow=c(1,1))

```



# Task 6. Create new information from satellite imagery: NDVI

Lets calculate the Normalized Difference Vegetation Index (NDVI) and see where the vegetation grows best:
Remember the formula for NDVI is: (NIR - RED) / (NIR + RED)

```{r}
# prep imagery
n3 <- raster(nov3) # RED
n4 <- raster(nov4) # NIR

ndvi <-  (n4 - n3) / (n4 + n3)
ndvi
plot(ndvi)

plot3D(dem, drape=ndvi)

### remove values below zero
ndvi[ndvi <= 0] <- NA
plot(ndvi)

## plot again
plot3D(dem, drape=ndvi)

## different way to plot in 2D
library(rasterVis)
levelplot(ndvi, col.regions=bpy.colors(100))

```

## ---- skip to task 8 if short on time

# Task 7: Create new information from satellite imagery: SAVI

```{r}
### another index SAVI (soil adjusted vegetation index)
ndvi <-  (n4 - n3) / (n4 + n3)
savi <-  (n4 - n3) / ((n4 + n3)*0.25) # with L=1 -> similar to NDVI

### let´s compare visually
par(mfrow=c(1,2))
plot(savi, main="SAVI");plot(ndvi, main="NDVI")
par(mfrow=c(1,1))


### Let´s make a new brick, including DEM and NDVI
data(nov2)
data(nov1)
n2 <- raster(nov2)
n1 <- raster(nov1)

myNewBrick <- brick(n4,n3,n2,n1,ndvi)
splom(myNewBrick)

```

# Task 8:  Unsupervised Classification - kmeans

```{r}
# run kmeans cluster analysis on image data to see the bands of growth better!
# read on Thresholding here:  https://rspatial.org/raster/rs/3-basicmath.html#vegetation-indices
ICE_df <- as.data.frame(myNewBrick)  
set.seed(99)
cluster_ICE <- kmeans(ICE_df, 4) ### kmeans, with 4 clusters
str(cluster_ICE)
# cluster_ICE <- cluster::clara(ICE_df, 4) ### clara, with 4 clusters

# convert cluster information into a raster for plotting
clusters <- raster(myNewBrick)   ## create an empty raster with same extent than ICE
clusters <- setValues(clusters, cluster_ICE$cluster) # convert cluster values into raster
clusters
plot(clusters)

# plot over the DEM
plot3D(dem, drape=clusters, col=c("red", "green", "blue", "yellow"))

# calculate the average spectral signature of 1-4 bands of growth
ICE_mean <- zonal(myNewBrick, clusters, fun="mean")  
ICE_mean  # see the values for ndvi (layer) being most distinct


```


# Task 9. What is the trend in de-/afforestation?  - Individual tree crown segmentation 
The ITC delineation approach finds local maxima within imagery, designates these as tree tops, then uses a decision tree method to grow individual crowns around the local maxima.
```{r}
library(itcSegment)

data(imgData) 
plot(imgData)

se<-itcIMG(imgData,epsg=32632)
summary(se)
plot(se,axes=T)

### let´s overlay them
plot(imgData)
plot(se,axes=T,add=T)
```


# Task 9: Visualise segmentation result in Leaflet

```{r}
### when we reproject the data to geographical coordinates...
se      # it is a Spatial object
crs(se) # what is its crs?
se.ll <- spTransform(se,CRS( "+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0 "))

### ...then we can combine them with leaflet

library(leaflet)
leaflet() %>% 
  addTiles() %>% 
  addRasterImage(imgData) %>% 
  addPolygons(data=se.ll, weight = 1, color = "black" )   # Tadaa
```



# Task 10:  a more demanding example! see content of folder...
```{r}
r <- raster("data/myDem_subset.tif")
r
plot(r)

r.se<-itcIMG(r,epsg=25829, ischm=T) ### slow on my laptop (solid 2-3mins)!!!
summary(r.se)
plot(r);plot(r.se,axes=T, add=TRUE)

# Adjust 'th' argument for excessive capture of small growth
r.se5<-itcIMG(r,epsg=25829,th = 5, ischm=T) # th - how low should algorithm be looking for canopy
plot(r);plot(r.se5,axes=T, add=TRUE)

# Write the result to shapefile
library(rgdal)
?writeOGR
td <- getwd()
writeOGR(r.se,td,"itcTrees_subset",driver="ESRI Shapefile" )


# want to see it in Leaflet?
library(sf)
rse <- st_read("itcTrees_subset.shp")
rse.ll <- st_transform(rse, crs = "+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0 ")

# Control question: where is this landscape from?
leaflet() %>% 
  addTiles() %>% 
  addProviderTiles("Esri.WorldPhysical") %>% 
 # addProviderTiles("Esri.WorldImagery") %>% 
  addRasterImage(projectRaster(r, crs = "+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0 ")) %>% 
  addPolygons(data=rse.ll, weight = 1, color = "black" )   # Neat :)
```



# The End.
Similar approach can be used to the mapping of socio-cultural phenomena, such as burial mounds in the landscape, growing urban sprawl, or tracing the outlines of scanned line drawings. (although in the latter two you may need to base the classification on reflectance or edge detection rather than elevation)

# Bibliography:
https://geoscripting-wur.github.io/AdvancedRasterAnalysis/
http://rspatial.org/spatial/rst/8-rastermanip.html
http://neondataskills.org/R/Image-Raster-Data-In-R/
https://geoscripting-wur.github.io/IntroToRaster/
http://wiki.landscapetoolbox.org/doku.php/remote_sensing_methods:home
https://rpubs.com/alobo/vectorOnraster