---
title: "Week 06 - Assessing Populations at Risk"
author: "Adela Sobotkova"
date: "16-February-2025 updated `r Sys.Date()`"
output:
  rmdformats::readthedown:
    highlight: kate
---

```{r setup, include=FALSE}
library(knitr)
library(rmdformats)
library(tidyverse)
library(sf)
library(mapboxapi)

## Global options
options(max.print="75")
opts_chunk$set(echo=TRUE,
	             cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```

## Task 3: Assessing Risk with Demographic data
This script continues on `05_NetworkDistancesDK.Rmd`. 
The task is to explore the most vulnerable parts of Midtjylland where 25%+ households don't have cars and are far from the nearest hospital.

### Carless Household data from DS
Number of households without cars are available via Danmark Statistik via the  "Familiernes bilrådighed (faktiske tal) efter rådighedsmønster, område" dataset. The "familier uden biler" sits in csv format as a hh_nocars.csv with the numbers and percentages corresponding to 2020 and 2021 respectively. Overall, there were 1178935 households without a car in 2020 and 1166668 in 2021. The number of course drops with time, while number of car-owning households go up steadily from 1,896,387 in 2020 to 1,929,511 in 2021

* Load municipality data 
* Create `no_cars_pct_mun` object by joining the `hh_nocar` object to the municipalities with `inner_join()`. Beware to preserve the spatial dimension of the `munic` for later use in a map. Look up the function to know how exactly to specify the shared columns. You may also want to rename the `pop` columns first so they are easier to map to `munic` 

```{r households-mun-nocars}
## YOUR CODE HERE
```

* In the next chunk of code we inspect the municipalities in a map to see if the names between the datasets overlap

* Create a thematic map of `no_cars_pct_mun` by filtering municipalities that have 25+ percent of households without a car. I recommend `mapview()`, but you are free to use your favorite library.  Mapview usually works well at the end of tidyverse pipeline but as it may interfere with other libraries, I recommend saving the script and upon restart, running only the `mapview()` code chunk so as to avoid interference. 

* *Which areas in Midtjylland have 25 and more percent of carless households?*
* If you notice any municipalities missing, reconcile their spelling in the `municipality` and `hh_nocars` objects.

```{r mapview-thematic-map}
## YOUR CODE HERE
```
Gosh, that was a ton of wrangling! But we are nearly there! :)


## Spatial overlay with sf

Spatial overlay is a very common operation when working with spatial data. It can be used to determine which features in one spatial layer overlap with another spatial layer, or extract data from a layer based on geographic information. In R, spatial overlay can be integrated directly into tidyverse-style data analysis pipelines using functions in sf. In our example, we want to determine the areas in Midtjylland with the greatest proportion of households without access to a car that also are beyond a 20 minute walk/cycling route from a hospital.

To do this, follow these steps:

* doublecheck that the coordinate reference system of the `no_car_pct_mun` dataset is 4326, the same CRS used by the isochrones;
* filter municipalities that fall within Midtjylland;
* extract only those municipalities with a percentage of households without cars of 25 percent or above;
* union your isochrones before the final step
* use the `st_difference()` function to “cut out” areas from those municipalities that overlap the 20-minute cycling isochrones.
* Once you complete this operation, visualize the result in your Mapbox map.


```{r risk-areas}
## YOUR CODE HERE

```


### Questions:

*1. Are there any areas that are located beyond a 20-minute bike-/carride of a hospital that have proportionally lower access to cars?* 

*2. Where are they and what is their aggregate area?*

*3. How would you ensure residents in these areas have reasonable access to hospitals?*

